<!-- To be compatible with workflowr, we need to specify the absolute directory under which the plots are generated, otherwise they will be saved into figure/<basename>.Rmd/<chunk-name>.<extension. and not visible in the html document: https://github.com/workflowr/workflowr/issues/111 -->
```{r template-options, include = FALSE}
library(ggbeeswarm)
knitr::opts_knit$set(
  base.dir = "docs"
  )
```

```{r set-gsea-collections}
ora_collections <- list(
  enricher = names(gmt_files_symbols),
  enrichGO = c("BP", "MF")
)

```

### Download results


<!-- Write ORA results to xlsx file -->
```{r write-ora_list-xlsx}
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'

file_xlsx_list <- list()
sr <- names(ora_collections)[1]
gc <- ora_collections[[sr]][1]
for(sr in names(ora_collections)) {
  file_xlsx_list[[sr]] <- list()
  for(gc in ora_collections[[sr]]) {
    # Generate file
    file_xlsx <- file.path('./docs/file', 
                           rmd_file, 
                           paste0('ora.', tolower(sr), '_', tolower(gc), '.xlsx')
                           )
    file_xlsx_list[[sr]][[gc]] <- file_xlsx
    dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)
    
    # Generate workbook
    wb <- createWorkbook()
    for(i in names(gse_list)) {
      if(gc != sr){
        for(direction in names(gse_list[[i]][[sr]])) {
          addWorksheet(wb, paste0(i,"-",direction))
          if(!is.null(gse_list[[i]][[sr]][[direction]][[gc]])) {
            res <- gse_list[[i]][[sr]][[direction]][[gc]]@result
            writeData(wb, paste0(i,"-",direction), res)
          }
        }
      }
      if(gc == sr){
        for(direction in names(gse_list[[i]][[sr]])) {
          addWorksheet(wb, paste0(i,"-",direction))
          if(!is.null(gse_list[[i]][[sr]][[direction]])) {
            res <- gse_list[[i]][[sr]][[direction]]@result
            writeData(wb, paste0(i,"-",direction), res)
          }
        }
      }
    }
    saveWorkbook(wb, file_xlsx, TRUE)
  }
}
```

<!-- Change gse_list names after writing to xlsx -->
```{r change-ora-names}
if(exists('comp_list_description')) {
  comp_names <- comp_list_description$comparison_long %>% 
    set_names(comp_list_description$comparison)
  comp_list_backup <- comp_list
  names_gse_list_backup <- names(gse_list)
  
  names(comp_list) <- comp_names[names(comp_list)]
  names(gse_list) <- comp_names[names(gse_list)]
}
```

<!-- Write links to download ORA results -->
```{r write-links-ora, echo = TRUE}
for(sr in names(file_xlsx_list)) {
  for(gc in names(file_xlsx_list[[sr]])) {
    file_xlsx <- gsub("docs/", "", file_xlsx_list[[sr]][[gc]])
    if(sr == 'enricher') {
      cat(paste0("- [**Over-enrichment analysis: ", gc, "**](", file_xlsx, ")"), '\n\n')
    }
    if(sr == 'ernrichGO') {
      cat(paste0("- [**Over-enrichment analysis: GO:", gc, "**](", file_xlsx, ")"), '\n\n')
    }
    if(sr == 'enrichKEGG') {
      cat(paste0("- [**Over-enrichment analysis: KEGG**](", file_xlsx, ")"), '\n\n')
    }
    if(sr == 'enrichReactome') {
      cat(paste0("- [**Over-enrichment analysis: Reactome**](", file_xlsx, ")"), '\n\n')
    }
  }
}
```




### Summary of results
```{r ora-summary-fdr}
res <- foreach(i = names(gse_list), .combine = rbind) %do% {
  foreach(sr = names(ora_collections), .combine = rbind) %do% {
    
    foreach(gc = ora_collections[[sr]], .combine = rbind) %do% {
      if(sr == 'enricher') {
        n_abs <- if(!is.null(gse_list[[i]][[sr]][['abs']][[gc]])) {
          gse_list[[i]][[sr]][['abs']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_down <- if(!is.null(gse_list[[i]][[sr]][['down']][[gc]])) {
          gse_list[[i]][[sr]][['down']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_up <- if(!is.null(gse_list[[i]][[sr]][['up']][[gc]])) {
          gse_list[[i]][[sr]][['up']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
      }
            if(sr == 'enrichGO') {
        n_abs <- if(!is.null(gse_list[[i]][[sr]][['abs']][[gc]])) {
          gse_list[[i]][[sr]][['abs']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_down <- if(!is.null(gse_list[[i]][[sr]][['down']][[gc]])) {
          gse_list[[i]][[sr]][['down']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_up <- if(!is.null(gse_list[[i]][[sr]][['up']][[gc]])) {
          gse_list[[i]][[sr]][['up']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
      }
        
        
        c(
          Comparison = i,
          `Collection` = gc,
          `Total` = n_abs,
          `Supressed` = n_down,
          `Activated` = n_up
        )
      
    }
  }
}

res <- res %>% data.frame(check.names = F)

```

#### Table
```{r ora-summary-fdr-table}
res %>% 
  data.frame(check.names = F) %>% 
  kbl(caption = 'Number of gene sets at FDR < 0.05',
              row.names = F) %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), 
              full_width = F) %>%
  column_spec(1, bold = TRUE)
```

#### Barplot
```{r gse-summary-fdr-barplot, fig.asp = 1.3}
res_plot <- res %>% 
  pivot_longer(cols = c('Total', 'Supressed', 'Activated'), names_to = 'Direction', values_to = 'Count') %>% 
  mutate(
    comb_names = paste(Comparison, Collection),
    Count = as.numeric(Count),
    Comparison = gsub("-over-", " over\n", Comparison),
    Comparison = gsub(" corrected", "\ncorrected", Comparison)
  ) %>% 
  ggplot(aes(Collection, Count, fill = Direction)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_grid(rows = vars(Comparison)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c(Supressed = '#74add1', 
                               Activated = '#f46d43',
                               Total = 'grey80')
  ) +
  labs(x = '', fill = '') +
  theme_facet +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_line(linewidth = one_pt/4, color = 'black'),
    strip.text = element_text(size = 5)
    
  )


subchunkify(
  print(res_plot), 
  fig.width = 3.5, 
  fig.height = 3.5*0.4,
  dpi = 600,
  prefix = 'gse-summary-fdr-barplot'
  # suffix = tolower(i) %>% gsub(" ", "-", .)
)
cat("\n\n")
```


### Dotplots

<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r ora-dotplot, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

use_linewidth <- 1/2.141959
root_section_number <- 6.3
section_n <- 1
subsection_n <- 1
n_top <- 10

for(i in names(gse_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h5> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h5>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(ora_collections)) {
    for(gc in ora_collections[[sr]]) {
      if(sr == 'enricher') {
        
        if(!is.null(gse_list[[i]][[sr]][["up"]][[gc]])) {
          x_up <- gse_list[[i]][[sr]][["up"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Up'
            ) %>% 
            head(n_top)
        } else {
          x_up <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["down"]][[gc]])) {
          x_down <- gse_list[[i]][[sr]][["down"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Down'
            ) %>% 
            head(n_top)
        } else {
          x_down <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["abs"]][[gc]])) {
          x_abs <- gse_list[[i]][[sr]][["abs"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05 & !(ID %in% x_up$ID) & !(ID %in% x_down$ID)) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Abs'
            )  %>% 
            head(n_top)
        } else {
          x_abs <- data.frame()
        }
        
        x <- rbind(x_up, x_down, x_abs)
      }
      
      if(sr == 'enrichGO') {
        
        if(!is.null(gse_list[[i]][[sr]][["up"]][[gc]])) {
          x_up <- gse_list[[i]][[sr]][["up"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Up'
            ) %>% 
            filter(simplify) %>% 
            head(n_top)
        } else {
          x_up <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["down"]][[gc]])) {
          x_down <- gse_list[[i]][[sr]][["down"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Down'
            ) %>% 
            filter(simplify) %>% 
            head(n_top)
        } else {
          x_down <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["abs"]][[gc]])) {
          x_abs <- gse_list[[i]][[sr]][["abs"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05 & !(ID %in% x_up$ID) & !(ID %in% x_down$ID)) %>% 
            filter(simplify) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Abs'
            )  %>% 
            head(n_top)
        } else {
          x_abs <- data.frame()
        }
        
        x <- rbind(x_up, x_down, x_abs)
      }
      
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h6> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h6>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
        
        x %<>% 
          mutate(
            log10P = -log10(p.adjust),
            Description = fct_reorder(Description, p.adjust, .desc = TRUE),
            Direction = factor(Direction, levels = c('Up', 'Down', 'Abs'))
            )
        
        res <- x %>% 
          ggplot(aes(x = log10P, y = Description, 
                     color = Direction,
                     size = GeneProp)) +
          geom_point(pch = 16, alpha = 0.8) + 
          facet_grid(rows = vars(Direction), scale = 'free_y', space = 'free_y') +
          scale_y_discrete(labels = default_labeller(50)) +
          scale_color_manual(values = c(Up = 'firebrick', Down = 'dodgerblue', Abs = 'grey50')) +
          scale_size_continuous(range=c(0.3, 2.5)) +
          scale_x_continuous(expand = expansion(mult = c(0.05, 0.05))) +
          labs(
            x = expression(paste("-lo", g[10],"(Adjusted P value)")),
            y = '',
            size = 'Gene proportion'
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = use_linewidth/2),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", linewidth = use_linewidth/2),
            axis.text.y = element_text(size = 4),
            axis.text.x = element_text(size = 4),
            
            legend.text = element_text(size = 4), 
            legend.title = element_text(size = 4),
            legend.key.width = unit(0.2, "cm")
          )
        
        subchunkify(res, 
                    fig.width = 3.5, 
                    fig.height = (3.5*0.3) + nrow(x)*0.10,
                    dpi = 600,
                    prefix = paste0('ora-dotplot'),
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        cat("\n\n")
      }
    }
  }
}

 cat("\n\n")
```

### Similarity matrix by gene set collection

```{r ora-jcmatrix-by-gset, echo=FALSE}
#
# Generate a list of JC similarity matrices by gene-set
#
library(enrichplot)
p_thrs <- 0.05
topn <- 50
use_comp <- names(gse_list)[1]
sr <- 'enricher'
jc_matrix_list <- foreach(use_comp = names(gse_list)) %do% {
  data_gse <- foreach(sr = names(ora_collections)) %do% {
    data_by_collection <- foreach(gc = ora_collections[[sr]])  %do% {
      
      if(sr == 'enricher') {
        res_direction <- foreach(direction = c('up', 'down', 'abs')) %do% {
          if(!is.null(gse_list[[use_comp]][[sr]][[direction]][[gc]])) {
            use_enrich_results <- gse_list[[use_comp]][[sr]][[direction]][[gc]]
            top_descriptions <- use_enrich_results@result %>% 
              filter(p.adjust <= p_thrs) %>% 
              head(topn) %>% 
              pull(Description)
            use_n_categories <-  use_enrich_results@result %>% 
              filter(p.adjust <= p_thrs) %>% 
              nrow
            
            if(length(top_descriptions) > 1) {
              use_enrich_results <- pairwise_termsim(
                use_enrich_results, 
                showCategory = use_n_categories, 
                method = "JC")
              use_enrich_results@termsim[top_descriptions, top_descriptions]
            } else {
              NULL
            }
          } else {
            NULL
          }
        }
        names(res_direction) <- c('up', 'down', 'abs')
        return(compact(res_direction))
      }
    }
    
    names(data_by_collection) <- ora_collections[[sr]]
    return(compact(data_by_collection))
  }
  names(data_gse) <- names(ora_collections)
  return(data_gse)
}
names(jc_matrix_list) <- names(gse_list)
```


Gene-set similarity matrix by comparison, gene set collections and direction. Heatmap showing the pair-wise similarity matrix of top `r topn` gene sets with an adjusted P value < `r p_thrs` in each gene group. Heatmap colors represent the Jaccard similarity coefficient.


```{r ora-jcmatrix-by-gset-heatmap, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )


library(NbClust)

root_section_number <- 6.5
section_n <- 1
subsection_n <- 1

col_fun <- colorRamp2(seq(0, 1, length.out = 4), brewer.pal(4, "GnBu"))

i <- names(jc_matrix_list)[1]
for(i in names(jc_matrix_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  
  cat("<h5> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h5>\n\n") #Caue results='asis',we must use raw HTML
  
  section_n <- section_n + 1
  
  for(sr in names(jc_matrix_list[[i]])) {
    for(gc in names(jc_matrix_list[[i]][[sr]])) {
      cat("<h6> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h6>\n\n") #Caue results='asis',we must use raw HTML
      
      for(direction in names(jc_matrix_list[[i]][[sr]][[gc]])) {
        use_mat <- jc_matrix_list[[i]][[sr]][[gc]][[direction]]
        if(nrow(use_mat) >= 10) {
          h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
          subsection_n <- subsection_n + 1
          colnames(use_mat) <- colnames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
          rownames(use_mat) <- rownames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
          
          # Define number of clusters
          use_max.nc <- ifelse(nrow(use_mat) <= 15, 5, round(nrow(use_mat) * 0.3))
          nbclust_res <- NbClust(data = NULL, diss = as.dist(1-use_mat), distance = NULL, 
                                 min.nc = 2, max.nc = use_max.nc, method = "ward.D",
                                 index = 'silhouette')
          
          
          n_split <- nbclust_res$Best.nc['Number_clusters']
          
          if(n_split == -Inf) {
            n_split <- NULL
          }
          # ha_top <- HeatmapAnnotation(
          #   foo = anno_block(
          #     # labels = c("Translation", "Cell division"),
          #     labels = 1:n_split,
          #     labels_gp = gpar(col = "black", fontsize = 8),
          #     gp = gpar(lwd = 0, lty = 0))
          # )
          
          ht <- Heatmap(
            use_mat,
            name = 'Jaccard index',
            column_title = paste('Direction:', direction),
            column_split = n_split,
            row_split = n_split,
            row_title = NULL,
            row_names_gp = gpar(fontsize = 4),
            col = col_fun,
            show_column_dend = FALSE,
            show_column_names = FALSE,
            border = TRUE,
            # top_annotation = ha_top,
            heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                                        labels_gp = gpar(fontsize = 8)
            ),
            width = unit(7, "cm"),
            
            clustering_method_rows = "ward.D",
            clustering_method_columns = "ward.D"
            # clustering_distance_rows = function(x) as.dist(1 - x),
            # clustering_distance_columns = function(x) as.dist(1 - x)
          )
          subchunkify(draw(ht),
                      fig.width = 7.2,
                      fig.height = (7.2*0.3) + nrow(use_mat)*0.15,
                      dpi = 600,
                      prefix = 'ora-jcmatrix-by-gset-heatmap',
                      suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc, '_', direction) )
          
          cat("\n\n")
        }
      }
    }
  }
}
```


<!-- Restore gse_list names after writing to xlsx -->
```{r restore-gse-names}
if(exists('comp_list_description')) {
  comp_list <- comp_list_backup
  names(gse_list) <- names_gse_list_backup
}
```
