<!-- To be compatible with workflowr, we need to specify the absolute directory under which the plots are generated, otherwise they will be saved into figure/<basename>.Rmd/<chunk-name>.<extension. and not visible in the html document: https://github.com/workflowr/workflowr/issues/111 -->
```{r template-options, include = FALSE}
library(ggbeeswarm)
knitr::opts_knit$set(
  base.dir = "docs"
  )
```

```{r set-gsea-collections}
gse_collections <- list(
  GSEA = names(gmt_files_symbols)
)

```

### Download results
<!-- Write GSE results to xlsx file -->
```{r write-gse_list-xlsx}
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'

file_xlsx_list <- list()
for(sr in names(gse_collections)) {
  file_xlsx_list[[sr]] <- list()
  for(gc in gse_collections[[sr]]) {
    # Generate file
    file_xlsx <- file.path('./docs/file', 
                           rmd_file, 
                           paste0('gse_', tolower(sr), '_', tolower(gc), '.xlsx')
                           )
    file_xlsx_list[[sr]][[gc]] <- file_xlsx
    dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)
    
    # Generate workbook
    wb <- createWorkbook()
    for(i in names(gse_list)) {
      addWorksheet(wb, i)
      if(gc != sr){
        if(!is.null(gse_list[[i]][[sr]][[gc]])) {
          res <- gse_list[[i]][[sr]][[gc]]@result
          writeData(wb, i, res)
        }
      }
      if(gc == sr){
        if(!is.null(gse_list[[i]][[sr]])) {
          res <- gse_list[[i]][[sr]]@result
          writeData(wb, i, res)
        }
      }
    }
    saveWorkbook(wb, file_xlsx, TRUE)
  }
}
```

<!-- Change gse_list names after writing to xlsx -->
```{r change-gse-names}
if(exists('comp_list_description')) {
  comp_names <- comp_list_description$comparison_long %>% 
    set_names(comp_list_description$comparison)
  
  comp_list_backup <- comp_list
  names_gse_list_backup <- names(gse_list)
  
  names(comp_list) <- comp_names[names(comp_list)]
  names(gse_list) <- comp_names[names(gse_list)]
}
```

<!-- Write links to download GSE results -->
```{r write-links-gse-2, echo = FALSE}
for(sr in names(file_xlsx_list)) {
  for(gc in names(file_xlsx_list[[sr]])) {
    file_xlsx <- gsub("docs/", "", file_xlsx_list[[sr]][[gc]])
    if(sr == 'GSEA') {
      cat(paste0("- [**Gene set enrichment analysis: ", gc, "**](", file_xlsx, ")"), '\n\n')
    }
  }
}
```


### Gene-set similarity matrix overall
```{r gse-jcmatrix-omni, echo=FALSE}
#
# Generate a list of JC similarity matrices
#

p_thrs <- 0.05
topn <- 20

use_comp <- names(gse_list)[1]
jc_matrix_list <- foreach(use_comp = names(gse_list)) %do% {
  data_gse <- foreach(sr = names(gse_collections)) %do% {
    data_by_collection <- foreach(gc = gse_collections[[sr]], .combine = rbind)  %do% {
      if(sr == 'GSEA') {
        x <- gse_list[[use_comp]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          arrange(p.adjust) %>%
          head(topn)
        # if(gc == 'msigdb.c2.cp') {
        #   x %<>% 
        #     filter(grepl('REACTOME|KEGG|WP', ID))
        # }
      }
      if(sr == 'gseGO') {
        x <- gse_list[[use_comp]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= p_thrs & simplify) %>% 
          arrange(p.adjust) %>%
          head(topn)
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        x <- gse_list[[use_comp]][[sr]]@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          arrange(p.adjust) %>%
          head(topn)
      }
      return(x)
    }
    return(data_by_collection)
  }

  data_gse <- purrr::compact(data_gse)
  gse_df <- bind_rows(data_gse)
  if(nrow(gse_df) > 1) {
    gse_df <- bind_rows(data_gse) %>% 
      mutate(
        Description = clean_msigdb_names(Description) %>% tolower() %>% str_to_title,
        Description = ifelse(grepl("^GO:", ID), paste0("GO:", Description), Description)
      )
    # # Select topn by group in selected gset
    # top_by_group <-   gse_df %>%
    #   # filter(group == use_comp) %>%
    #   # filter(p.adjust < p_thrs) %>%
    #   arrange(p.adjust) %>% pull(Description) %>%
    #   head(topn)
    
    # Data frame to plot
    res_top <- gse_df %>%
      # filter(Description %in% top_by_group) %>%
      mutate(
        log.p.adjust = -log10(p.adjust),
        Description = factor(Description)
      )
    
    # Calculate jaccard similarity matrix
    gset_core_enrichment <- foreach(i = levels(res_top$Description)) %do% {
      res_top %>% filter(Description == i) %>% pull(core_enrichment) %>%
        paste(collapse = '/') %>%
        strsplit('/') %>% unlist %>% unique
    }
    names(gset_core_enrichment) <- levels(res_top$Description)
    
    use_terms <- res_top$Description %>% unique
    jc_matrix <- pairwise_terms_jc(gset_core_enrichment[use_terms])
  } else {
    jc_matrix <- NULL
  }  
  return(jc_matrix)
}

names(jc_matrix_list) <- names(gse_list)
jc_matrix_list <- purrr::compact(jc_matrix_list)

```

```{r gse-jcmatrix-by-gset, echo=FALSE}
#
# Generate a list of JC similarity matrices by gene-set
#
library(enrichplot)
p_thrs <- 0.05
topn <- 50
use_comp <- names(gse_list)[1]
jc_matrix_list <- foreach(use_comp = names(gse_list)) %do% {
  data_gse <- foreach(sr = names(gse_collections)) %do% {
    data_by_collection <- foreach(gc = gse_collections[[sr]])  %do% {
      
      if(sr == 'GSEA') {
        use_enrich_results <- gse_list[[use_comp]][[sr]][[gc]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      if(sr == 'gseGO') {
        use_enrich_results <- gse_list[[use_comp]][[sr]][[gc]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs & simplify) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        use_enrich_results <- gse_list[[use_comp]][[sr]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      
      if(length(top_descriptions) > 1) {
        use_enrich_results <- pairwise_termsim(
          use_enrich_results, 
          showCategory = use_n_categories, 
          method = "JC")
        use_enrich_results@termsim[top_descriptions, top_descriptions]
      } else {
        NULL
      }
    }
    names(data_by_collection) <- gse_collections[[sr]]
    return(compact(data_by_collection))
  }
  names(data_gse) <- names(gse_collections)
  return(data_gse)
}
names(jc_matrix_list) <- names(gse_list)
```


Gene-set similarity matrix by comparison and gene set collections. Heatmap showing the pair-wise similarity matrix of top `r topn` gene sets with an adjusted P value < `r p_thrs` in each gene group. Heatmap colors represent the Jaccard similarity coefficient.

```{r gse-jcmatrix-by-gset-heatmap, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

library(NbClust)

root_section_number <- 6.5
section_n <- 1
subsection_n <- 1

col_fun <- colorRamp2(seq(0, 1, length.out = 4), brewer.pal(4, "GnBu"))

i <- names(jc_matrix_list)[1]
for(i in names(jc_matrix_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  
  section_n <- section_n + 1
  
  for(sr in names(jc_matrix_list[[i]])) {
    
    for(gc in names(jc_matrix_list[[i]][[sr]])) {
      use_mat <- jc_matrix_list[[i]][[sr]][[gc]]
      if(nrow(use_mat) >= 10) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
        colnames(use_mat) <- colnames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
        rownames(use_mat) <- rownames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
        
        # Define number of clusters
        use_max.nc <- ifelse(nrow(use_mat) <= 15, 5, round(nrow(use_mat) * 0.3))
        nbclust_res <- NbClust(data = NULL, diss = as.dist(1-use_mat), distance = NULL, 
                               min.nc = 2, max.nc = use_max.nc, method = "ward.D",
                               index = 'silhouette')
        
        
        n_split <- nbclust_res$Best.nc['Number_clusters']
        
        # ha_top <- HeatmapAnnotation(
        #   foo = anno_block(
        #     # labels = c("Translation", "Cell division"),
        #     labels = 1:n_split,
        #     labels_gp = gpar(col = "black", fontsize = 8),
        #     gp = gpar(lwd = 0, lty = 0))
        # )
        
        ht <- Heatmap(
          use_mat,
          name = 'Jaccard index',
          column_split = n_split,
          row_split = n_split,
          column_title = NULL,
          row_title = NULL,
          row_names_gp = gpar(fontsize = 4),
          col = col_fun,
          show_column_dend = FALSE,
          show_column_names = FALSE,
          border = TRUE,
          # top_annotation = ha_top,
          heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                                      labels_gp = gpar(fontsize = 8)
          ),
          width = unit(7, "cm"),
          
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D"
          # clustering_distance_rows = function(x) as.dist(1 - x),
          # clustering_distance_columns = function(x) as.dist(1 - x)
        )
        
        subchunkify(draw(ht),
                    fig.width = 7.2,
                    fig.height = (7.2*0.3) + nrow(use_mat)*0.15,
                    dpi = 600,
                    prefix = 'gse-jcmatrix-by-gset-heatmap',
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        cat("\n\n")
      }
    }
  }
}
```



<!-- Restore gse_list names after writing to xlsx -->
```{r restore-gse-names}
if(exists('comp_list_description')) {
  comp_list <- comp_list_backup
  names(gse_list) <- names_gse_list_backup
}
```


### Future: Get clusters and represent Dotplots by clusters (see yuwei/tumor-fasting)
