---
title: Analysis of CTCs from HT29_Li.M2 model
subtitle: Explorative analysis of human reads removing PC1
author: Francesc Castro-Giner
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  se_path: ./data/rnaseq/HT29_Li.M2/se_hs.rds
  output_dir: ./output/rnaseq/HT29_Li.M2/exploratory
  group: anatomical_origin
  min_counts: 5
  min_present_prop: 0.70
  msigdb_dir: ./data/resources/MSigDB
  msigdb_vs: v2026.1.Hs
  ann_vars: anatomical_origin,sample_type,n_cells,detected,sum,subsets_Mito_percent,subsets_Ribo_percent,mouse_id,cc_phase,cc_g1_score_cy,cc_s_score_cy,cc_g2m_score_cy
  run_cl_mk: FALSE
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"), fig.width = 3.5, fig.asp = 0.618)

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)
library(arsenal)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(circlize)
library(GSVA)
library(babelgene)
library(dynamicTreeCut)
library(dendextend)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
source('./code/R-functions/pairwise_terms_jc.r')
source('./code/R-functions/eigencorval.R')



clean_msigdb_names <- function(x) x %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Define additional functions
```{r add-functions}
clean_sample_alias <- function(x) 
  gsub('o[0-9]*_[0-9]*-|_gDNA.*', '', x) %>% gsub(".*-MS", "MS", .) %>% gsub("_S[0-9]+_R[0-9].*", "", .) %>% gsub('__STARpass1', '', .) %>% gsub(".k2report", "", .) %>% gsub(".*-|.disambiguated.*", "", .)
```

Load MSigDB gene sets
```{r load-msigdb}
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/c2.cp.', params$msigdb_vs, '.symbols.gmt')
)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```


## Data wrangling

### Human data
Load SE object
```{r dw-load-se-hs}
se <- readRDS(params$se_path)
```

Cell cycle annotation with cyclone, before filtering low abundance genes: *Users should not filter out low-abundance genes before applying cyclone. Even if a gene is not expressed in any cell, it may still be useful for classification if it is phase-specific.*
```{r dw-cc-cyclone}
cc.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", 
    package="scran"))
res <- cyclone(se, cc.pairs)
se$cc_phase_cy <- res$phases
se$cc_g1_score_cy <- res$normalized.scores$G1
se$cc_s_score_cy <- res$normalized.scores$S
se$cc_g2m_score_cy <- res$normalized.scores$G2M
cc_assignments_cy <- res
```


Filtering out low-abundance features. Keeping those features with at least 5 counts in at least 50% of samples of the smallest group.
```{r dw-subset-feature-abundance-hs}
# Not done: it is done at DGE level, for each comparison
use_rows <- filterByExpr(se, group=colData(se)[[params$group]], min.count = 5, min.total.count = 15, min.prop = 0.5)
se <- se[use_rows,]
```

TMM normalization
```{r dw-tmm-hs}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)
```

CPM calculation
```{r dw-cpm-hs}
assay(se, 'cpm') <- edgeR::cpm(assay(se, 'normcounts'))
assay(se, 'logcpm') <- edgeR::cpm(assay(se, 'normcounts'), log = TRUE)
```



### Additional objects
Selection of HVG
```{r conf-hvg}
# Variance modelling
fit_res <- modelGeneVar(se)
fit_md <- metadata(fit_res)
# Define highly variable genes
hvg <- getTopHVGs(fit_res, prop = 0.1)
```

Annotation variables and colors
```{r conf-annot}
ann_vars <- strsplit(params$ann_vars, ",") %>% unlist %>% 
  intersect(., colData(se) %>% colnames)
ann_df <- colData(se) %>% data.frame %>% 
  dplyr::select(all_of(ann_vars))

ann_colors <- foreach(i = ann_vars) %do% {
  get_palette(i)
}
names(ann_colors) <- ann_vars
ann_colors <- compact(ann_colors)
```


## PCA
```{r pca-conf}
use_rows <- hvg
mat <- assay(se[use_rows,], 'logcounts')
rownames(mat) <- make.names(rowData(se[use_rows,])$gene_name,unique = T)
# Create annotation, keeping only variables with more than 1 unique value
use_cols <- ann_df %>% dplyr::select(where(~n_distinct(.) > 1)) %>% colnames
p <- PCAtools::pca(mat, metadata = colData(se) %>% data.frame %>% dplyr::select(all_of(use_cols)))
```


### Determine optimum number of PCs to retain
Horn’s parallel analysis and elbow method
```{r pca-elbow, fig.width = 3.1, fig.asp = 0.9}
horn <- parallelPCA(mat)
elbow <- findElbowPoint(p$variance)
# horn$n
# elbow
screeplot(
  p,
  axisLabSize = 8, titleLabSize = 8,
  components = getComponents(p, 1:20),
  vline = c(horn$n, elbow)
) +
  geom_label(aes(x = horn$n + 0.5, y = 25, label = 'Horn\'s', vjust = -1),
             size = geom_text_size) +
  geom_label(aes(x = elbow + 0.5, y = 50, label = 'Elbow method', vjust = -1),
             size = geom_text_size)
```



### Variable correlation
Plot showing the correlation of PCs with sample-level variables. 
```{r pca-eigencor-plot, fig.width = 7.2}
use_rows <- hvg
n_pc <- seq_len(horn$n)
if(horn$n <= 2)
  n_pc <- 1:3

eigencor_res <- eigencorval(p, components = getComponents(p, n_pc), metavars = colnames(p$metadata))
```

```{r pca-eigencor-plot-curated, fig.width = 7.2}
use_rows <- hvg
n_pc <- seq_len(horn$n)
if(horn$n <= 2)
  n_pc <- 1:3

selected_vars <- c(
  `Anatomical origin` = 'anatomical_origin',
  `CTC type` = 'sample_type',
  `Number of cells` = 'n_cells',
  `Detected genes` = 'detected',
  `Total counts` = 'sum',
  `Mitochondrial %` = 'subsets_Mito_percent',
  `Ribosomal %` = 'subsets_Ribo_percent',
  `Mouse ID` = 'mouse_id',
  `G1 score` = 'cc_g1_score_cy',
  `S score` = 'cc_s_score_cy',
  `G2M score` = 'cc_g2m_score_cy'
)

p_curated <- p  
p_curated$metadata <- p_curated$metadata %>% 
  dplyr::rename(all_of(selected_vars))
eigencor_res <- eigencorval(p_curated, components = getComponents(p_curated, n_pc), metavars = names(selected_vars))

# Plot correlation
eigencorplot(p_curated,
             components = getComponents(p_curated, n_pc),
             metavars = names(selected_vars),
             fontCorval = 2,
             cexCorval = 0.8,
             main = 'PC sample variables correlation')


```

## Hierarchical clustering removing PC1
```{r hclust-conf}
dst <- dist(p$rotated[,2:10], method = 'euclidean')
hc <-  dst %>% hclust(method = 'ward.D2')
dend <- hc %>% as.dendrogram()
column_dend<- dend

## Find Clusters
clusters_original <- cutreeDynamic(hc, distM = as.matrix(dst),
                                   minClusterSize = 15, method="tree")
if(min(clusters_original) == 0) {
  clusters_original <- clusters_original + 1
}

## Add clustering info to se


cluster_info <- data.frame(
  sample_alias = rownames(p$rotated)[order.dendrogram(dend)],
  cluster_rmpc1 = as.character(clusters_original[order.dendrogram(dend)]) %>% 
    factor(levels = as.character(unique(clusters_original)))
  )

if(!is.null(se$cluster_rmpc1))
  se$cluster_rmpc1 <- NULL
colData(se) <- colData(se) %>% data.frame %>% 
  left_join(cluster_info) %>% 
  mutate(torownames = sample_alias) %>%
  column_to_rownames('torownames') %>% 
  DataFrame

ann_df$cluster_rmpc1 <- colData(se)[rownames(ann_df), 'cluster_rmpc1'] %>% as.character
```

### Dendrogram
Plot hierarchical clustering
```{r hclust-plot, fig.width=7.2}
## Sort clusters to the order of the dendrogram:
clusters <- clusters_original[order.dendrogram(dend)]
clusters_numbers <- unique(clusters) - (0 %in% clusters)
n_clusters <- length(clusters_numbers)
cluster_cols <- rainbow_hcl(n_clusters) %>% 
  set_names(1:n_clusters)
column_dend_colored <- dend %>% 
         branches_attr_by_clusters(clusters, values = cluster_cols) 


# Add color lables
bar_var_sorted <- ann_df[[params$group]][order.dendrogram(dend)]
bar_color <- get_palette(params$group)[bar_var_sorted]

# Plot colored dendrogram
plot(column_dend_colored)
colored_bars(colors = bar_color, dend = column_dend_colored, rowLabels = params$group)

# Add cluster color to annotation df
ann_colors$cluster_rmpc1 <- cluster_cols
```


### Heatmap simplified
```{r hclust-heatmap-simplified, fig.width=4.5, fig.asp=0.8}
mat <- p$rotated[,2:10] %>% t

pc_var <- paste0(p$variance[rownames(mat)] %>% round(1), '%')
rownames(mat) <- paste(rownames(mat), pc_var, sep = ' (') %>% paste0(')')

zmat <- t(scale(t(mat)))

use_annot <- colData(se) %>% data.frame %>% 
  mutate(
    anatomical_origin = ifelse(anatomical_origin == 'heart_puncture', 'heart puncture', anatomical_origin),
    anatomical_origin = ifelse(anatomical_origin == 'portal_vein', 'portal vein', anatomical_origin)
    )

# Column annotation
heatmap_col_annot_vars <- c('anatomical_origin')

column_ha_colors <- lapply(heatmap_col_annot_vars, get_palette) %>% 
    set_names(heatmap_col_annot_vars) %>% 
    compact

column_ha <-  HeatmapAnnotation(
    df = use_annot %>% dplyr::select(all_of(heatmap_col_annot_vars)),
    col = column_ha_colors,
    annotation_name_gp = gpar(fontsize = 8),
    simple_anno_size = unit(0.2, "cm"),
    annotation_legend_param = list(
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 8)
    )
  )

ht <- Heatmap(
  mat,
  name = 'Loadings',
  row_names_gp = gpar(fontsize = 8),
  show_column_names = FALSE,
  show_column_dend = TRUE,
  cluster_rows = FALSE,
  # cluster_columns = hc,
  
  top_annotation = column_ha,
  
  column_split = as.character(use_annot$cluster_rmpc1),
  column_title = "Cluster %s",
  column_title_gp = gpar(fontsize = 8),
  
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 8),
    labels_gp = gpar(fontsize = 8),
    ncol = 2
  )
)

draw(
  ht, 
  annotation_legend_side = "bottom"
)

cat("\n\n")

```





### Descriptive table
```{r hclust-descriptive-table}
use_formula <- as.formula(paste0(params$group, " ~ cluster_rmpc1"))
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE)
summary(tab1)

# Cochran–Armitage trend test
use_table <- table(colData(se)[[params$group]], se$cluster_rmpc1)

res <- prop.trend.test(
  x = use_table['portal_vein', ], 
  n = table(se$cluster_rmpc1)
)

broom::tidy(res) %>% 
  mutate(one.sided.p = p.value / 2) %>% 
  kbl(caption = 'Cochran–Armitage trend test') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
#

tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE, strata=sample_type)
summary(tab1)

use_formula <- as.formula(paste0("cluster_rmpc1", "~", paste(ann_vars, collapse = "+")))
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE)
summary(tab1)
```


```{r hclust-12grouped-descriptive-table}
se$cluster_rmpc1_cl3 <- ifelse(as.character(se$cluster_rmpc1) %in% c('1', '2'), '1-2', '3')
use_formula <- as.formula(paste0(params$group, " ~ cluster_rmpc1_cl3"))
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE)
summary(tab1, pfootnote=TRUE)



tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE, strata=sample_type)
summary(tab1, pfootnote=TRUE)

use_formula <- as.formula(paste0("cluster_rmpc1_cl3", "~", paste(ann_vars, collapse = "+")))
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE)
summary(tab1, pfootnote=TRUE)

```



## Cluster markers: Differential expression
```{r hclust-mk-conf}
clvar <- 'cluster_rmpc1'
se$uclust <- colData(se)[[clvar]]

# Define comparisons
x <- "1"
comp_list <- foreach(x = levels(se$uclust)) %do%{
  y <- setdiff( levels(se$uclust), x)
  list(
    colnames(se)[as.character(se$uclust) == x],
    colnames(se)[as.character(se$uclust)%in% y]
  ) %>% set_names(c(x, paste(y, collapse = "-")))
}
names(comp_list) <- levels(se$uclust)

# RUN edgeR_dge and gse_omnibus
if(params$run_cl_mk){
  # Get ensembl annotation
  ensembl <-  biomaRt::useEnsembl(biomart="genes", dataset="hsapiens_gene_ensembl",
                                  mirror = 'useast')
  gene_desc <- biomaRt::getBM(attributes=c('external_gene_name','description'),
                              filters = 'external_gene_name',
                              values = rowData(se)$gene_name, mart =ensembl) %>%
    dplyr::rename('gene_name' = 'external_gene_name') %>%
    unique
  
  gene_desc <- gene_desc[!duplicated(gene_desc$gene_name),]
  # saveRDS(gene_desc, file = file.path(dirname(params$se_path), 'ensembl_gene_desc.rds'))
  
  # Run differential expression analysis using EdgeR QLF
  use_comp <-  comp_list[[1]]
  dge_list <- foreach(use_comp = comp_list) %do%{
    se_cols <- se$sample_alias %in% unlist(use_comp)
    use_se <- se[,se_cols]
    use_se$group <- ifelse(
      use_se$sample_alias %in% use_comp[[1]],
      names(use_comp)[1],
      names(use_comp)[2]
    )
    
    if(is.null(use_comp$batch)) {
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        design_formula = "~ group",
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
      )
    } else {
      design_formula = paste0("~ ", use_comp$batch," + group")
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        batch_vars = use_comp$batch,
        design_formula = design_formula,
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
      )
    }
    
    # Add gene description
    use_res <- dge$results %>%  left_join(., gene_desc, relationship = "many-to-many")
    dge$results <- use_res %>% 
      filter(!duplicated(feature)) %>% 
      mutate(rownames = feature) %>% 
      column_to_rownames('rownames')
    
    return(dge)
  }
  names(dge_list) <- names(comp_list)
  saveRDS(dge_list, file = file.path(params$output_dir, paste0('dge-',clvar,'.rds')))
}

dge_list <- readRDS(file.path(params$output_dir, paste0('dge-',clvar,'.rds')))
```

<!-- Run template -->
```{r knitr-template-rnaseq-deg-edger}
# Configuration: variable evalauted in differential expression is by default included
heatmap_col_annot_vars <- c('mouse_id', 'sample_type')

# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-deg-edger.Rmd")
  )
```
`r knit(text = src)`

## Cluster markers : ORA
```{r hclust-mk-ora-conf}
# Run ORA gse_omnibus
if(params$run_cl_mk){
  
  # Run functional enrichment
  dge <-  dge_list[[2]]
  gse_list <- foreach(dge = dge_list) %do%{
    gse_res <- gse_omnibus(
      feature_names = dge$results$gene_name,
      p = dge$results$FDR,
      fc = dge$results$logFC,
      fc_thrs = 0.5,
      return_sets = c('up', 'down'),
      gmt_files = gmt_files_symbols,
      
      run_enricher = TRUE,
      run_enrichGO = FALSE,
      run_enrichKEGG = FALSE,
      run_enrichReactome = FALSE,
      
      run_GSEA = FALSE,
      run_gseGO = FALSE,
      run_gseKEGG = FALSE,
      run_gseReactome = FALSE,
      
      go_simplify_p = 0.05,
      args_gse = list(minGSSize = 10, 
                      maxGSSize = 250, 
                      pvalueCutoff = 1,
                      BPPARAM = BiocParallel::MulticoreParam(1)
      ),
      args_enricher = list(minGSSize = 10, 
                           maxGSSize = 250, 
                           pvalueCutoff = 1, 
                           qvalueCutoff = 1
      )
      
    )
    
    return(gse_res)
    
  }
  names(gse_list) <- names(dge_list)
  saveRDS(gse_list, file = file.path(params$output_dir,paste0('gse_omnibus_ora-',clvar,'.rds')))
}
gse_list <- readRDS(file.path(params$output_dir, paste0('gse_omnibus_ora-',clvar,'.rds')))
ora_collections <- list(
  enricher = names(gmt_files_symbols)
)

``` 



### Download results


<!-- Write ORA results to xlsx file -->
```{r write-ora_list-xlsx}
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'

file_xlsx_list <- list()
sr <- names(ora_collections)[1]
gc <- ora_collections[[sr]][1]
for(sr in names(ora_collections)) {
  file_xlsx_list[[sr]] <- list()
  for(gc in ora_collections[[sr]]) {
    # Generate file
    file_xlsx <- file.path('./docs/file', 
                           rmd_file, 
                           paste0('ora.', tolower(sr), '_', tolower(gc), '.xlsx')
                           )
    file_xlsx_list[[sr]][[gc]] <- file_xlsx
    dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)
    
    # Generate workbook
    wb <- createWorkbook()
    for(i in names(gse_list)) {
      if(gc != sr){
        for(direction in names(gse_list[[i]][[sr]])) {
          addWorksheet(wb, paste0(i,"-",direction))
          if(!is.null(gse_list[[i]][[sr]][[direction]][[gc]])) {
            res <- gse_list[[i]][[sr]][[direction]][[gc]]@result
            writeData(wb, paste0(i,"-",direction), res)
          }
        }
      }
      if(gc == sr){
        for(direction in names(gse_list[[i]][[sr]])) {
          addWorksheet(wb, paste0(i,"-",direction))
          if(!is.null(gse_list[[i]][[sr]][[direction]])) {
            res <- gse_list[[i]][[sr]][[direction]]@result
            writeData(wb, paste0(i,"-",direction), res)
          }
        }
      }
    }
    saveWorkbook(wb, file_xlsx, TRUE)
  }
}
```

<!-- Change gse_list names after writing to xlsx -->
```{r change-ora-names}
if(exists('comp_list_description')) {
  comp_names <- comp_list_description$comparison_long %>% 
    set_names(comp_list_description$comparison)
  comp_list_backup <- comp_list
  names_gse_list_backup <- names(gse_list)
  
  names(comp_list) <- comp_names[names(comp_list)]
  names(gse_list) <- comp_names[names(gse_list)]
}
```

<!-- Write links to download ORA results -->
```{r write-links-ora, echo = TRUE}
for(sr in names(file_xlsx_list)) {
  for(gc in names(file_xlsx_list[[sr]])) {
    file_xlsx <- gsub("docs/", "", file_xlsx_list[[sr]][[gc]])
    if(sr == 'enricher') {
      cat(paste0("- [**Over-enrichment analysis: ", gc, "**](", file_xlsx, ")"), '\n\n')
    }
  }
}
```




### Summary of results
```{r ora-summary-fdr}
res <- foreach(i = names(gse_list), .combine = rbind) %do% {
  foreach(sr = names(ora_collections), .combine = rbind) %do% {
    
    foreach(gc = ora_collections[[sr]], .combine = rbind) %do% {
      if(sr == 'enricher') {
        n_abs <- if(!is.null(gse_list[[i]][[sr]][['abs']][[gc]])) {
          gse_list[[i]][[sr]][['abs']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_down <- if(!is.null(gse_list[[i]][[sr]][['down']][[gc]])) {
          gse_list[[i]][[sr]][['down']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_up <- if(!is.null(gse_list[[i]][[sr]][['up']][[gc]])) {
          gse_list[[i]][[sr]][['up']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
      }
            if(sr == 'enrichGO') {
        n_abs <- if(!is.null(gse_list[[i]][[sr]][['abs']][[gc]])) {
          gse_list[[i]][[sr]][['abs']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_down <- if(!is.null(gse_list[[i]][[sr]][['down']][[gc]])) {
          gse_list[[i]][[sr]][['down']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
        n_up <- if(!is.null(gse_list[[i]][[sr]][['up']][[gc]])) {
          gse_list[[i]][[sr]][['up']][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% nrow
        } else {
          0
        }
      }
        
        
        c(
          Comparison = i,
          `Collection` = gc,
          `Total` = n_abs,
          `Supressed` = n_down,
          `Activated` = n_up
        )
      
    }
  }
}

res <- res %>% data.frame(check.names = F)

```

#### Table
```{r ora-summary-fdr-table}
res %>% 
  data.frame(check.names = F) %>% 
  kbl(caption = 'Number of gene sets at FDR < 0.05',
              row.names = F) %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), 
              full_width = F) %>%
  column_spec(1, bold = TRUE)
```

#### Barplot
```{r gse-summary-fdr-barplot, fig.asp = 1.3}
res_plot <- res %>% 
  pivot_longer(cols = c('Total', 'Supressed', 'Activated'), names_to = 'Direction', values_to = 'Count') %>% 
  mutate(
    comb_names = paste(Comparison, Collection),
    Count = as.numeric(Count),
    Comparison = gsub("-over-", " over\n", Comparison),
    Comparison = gsub(" corrected", "\ncorrected", Comparison)
  ) %>% 
  ggplot(aes(Collection, Count, fill = Direction)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_grid(rows = vars(Comparison)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c(Supressed = '#74add1', 
                               Activated = '#f46d43',
                               Total = 'grey80')
  ) +
  labs(x = '', fill = '') +
  theme_facet +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_line(linewidth = one_pt/4, color = 'black'),
    strip.text = element_text(size = 5)
    
  )


subchunkify(
  print(res_plot), 
  fig.width = 3.5, 
  fig.height = 3.5*0.4,
  dpi = 600,
  prefix = 'gse-summary-fdr-barplot'
  # suffix = tolower(i) %>% gsub(" ", "-", .)
)
cat("\n\n")
```


### Dotplots

<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r ora-dotplot, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

use_linewidth <- 1/2.141959
root_section_number <- 6.3
section_n <- 1
subsection_n <- 1
n_top <- 10

for(i in names(gse_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h5> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h5>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(ora_collections)) {
    for(gc in ora_collections[[sr]]) {
      if(sr == 'enricher') {
        
        if(!is.null(gse_list[[i]][[sr]][["up"]][[gc]])) {
          x_up <- gse_list[[i]][[sr]][["up"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Up'
            ) %>% 
            head(n_top)
        } else {
          x_up <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["down"]][[gc]])) {
          x_down <- gse_list[[i]][[sr]][["down"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Down'
            ) %>% 
            head(n_top)
        } else {
          x_down <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["abs"]][[gc]])) {
          x_abs <- gse_list[[i]][[sr]][["abs"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05 & !(ID %in% x_up$ID) & !(ID %in% x_down$ID)) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Abs'
            )  %>% 
            head(n_top)
        } else {
          x_abs <- data.frame()
        }
        
        x <- rbind(x_up, x_down, x_abs)
      }
      
      if(sr == 'enrichGO') {
        
        if(!is.null(gse_list[[i]][[sr]][["up"]][[gc]])) {
          x_up <- gse_list[[i]][[sr]][["up"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Up'
            ) %>% 
            filter(simplify) %>% 
            head(n_top)
        } else {
          x_up <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["down"]][[gc]])) {
          x_down <- gse_list[[i]][[sr]][["down"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Down'
            ) %>% 
            filter(simplify) %>% 
            head(n_top)
        } else {
          x_down <- data.frame()
        }
        if(!is.null(gse_list[[i]][[sr]][["abs"]][[gc]])) {
          x_abs <- gse_list[[i]][[sr]][["abs"]][[gc]]@result %>% 
            filter(p.adjust <= 0.05 & !(ID %in% x_up$ID) & !(ID %in% x_down$ID)) %>% 
            filter(simplify) %>% 
            mutate(
              Description = clean_msigdb_names(Description),
              Direction = 'Abs'
            )  %>% 
            head(n_top)
        } else {
          x_abs <- data.frame()
        }
        
        x <- rbind(x_up, x_down, x_abs)
      }
      
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h6> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h6>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
        
        x %<>% 
          mutate(
            log10P = -log10(p.adjust),
            Description = fct_reorder(Description, p.adjust, .desc = TRUE),
            Direction = factor(Direction, levels = c('Up', 'Down', 'Abs'))
            )
        
        res <- x %>% 
          ggplot(aes(x = log10P, y = Description, 
                     color = Direction,
                     size = GeneProp)) +
          geom_point(pch = 16, alpha = 0.8) + 
          facet_grid(rows = vars(Direction), scale = 'free_y', space = 'free_y') +
          scale_y_discrete(labels = default_labeller(50)) +
          scale_color_manual(values = c(Up = 'firebrick', Down = 'dodgerblue', Abs = 'grey50')) +
          scale_size_continuous(range=c(0.3, 2.5)) +
          scale_x_continuous(expand = expansion(mult = c(0.05, 0.05))) +
          labs(
            x = expression(paste("-lo", g[10],"(Adjusted P value)")),
            y = '',
            size = 'Gene proportion'
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = use_linewidth/2),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", linewidth = use_linewidth/2),
            axis.text.y = element_text(size = 4),
            axis.text.x = element_text(size = 4),
            
            legend.text = element_text(size = 4), 
            legend.title = element_text(size = 4),
            legend.key.width = unit(0.2, "cm")
          )
        
        subchunkify(res, 
                    fig.width = 3.5, 
                    fig.height = (3.5*0.3) + nrow(x)*0.10,
                    dpi = 600,
                    prefix = paste0('ora-dotplot'),
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        cat("\n\n")
      }
    }
  }
}

 cat("\n\n")
```


