---
title: Analysis post-mortem samples from patient 031
subtitle: Explorative analysis of human reads
author: Francesc Castro-Giner
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  se_path: ./data/rnaseq/patient/se.rds
  output_dir: ./output/rnaseq/patient/exploratory
  group: anatomical_origin_grouped
  min_counts: 5
  min_present_prop: 0.70
  msigdb_dir: ./data/resources/MSigDB
  msigdb_vs: v2026.1.Hs
  ann_vars: anatomical_origin,anatomical_origin_grouped,sample_type,n_cells,detected,sum,subsets_Mito_percent,subsets_Ribo_percent,cc_phase_cy,cc_g1_score_cy,cc_s_score_cy,cc_g2m_score_cy
  run_cl_mk: FALSE
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"), fig.width = 3.5, fig.asp = 0.618)

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)
library(arsenal)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(circlize)
library(GSVA)
library(babelgene)
library(dynamicTreeCut)
library(dendextend)

library(GSEABase)
library(PerformanceAnalytics)
library(igraph)
library(rstatix)
library(ggpubr)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
source('./code/R-functions/pairwise_terms_jc.r')
source('./code/R-functions/eigencorval.R')



clean_msigdb_names <- function(x) x %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Define additional functions
```{r add-functions}
clean_sample_alias <- function(x) 
  gsub('o[0-9]*_[0-9]*-|_gDNA.*', '', x) %>% gsub(".*-MS", "MS", .) %>% gsub("_S[0-9]+_R[0-9].*", "", .) %>% gsub('__STARpass1', '', .) %>% gsub(".k2report", "", .) %>% gsub(".*-|.disambiguated.*", "", .)
```

Load MSigDB gene sets
```{r load-msigdb}
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/c2.cp.', params$msigdb_vs, '.symbols.gmt')
)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```


## Data wrangling

### Human data
Load SE object
```{r dw-load-se-hs}
se <- readRDS(params$se_path)

# Rename anatomical origin
se$anatomical_origin_grouped <- case_when(
  se$anatomical_origin == "inferior_vena_cava_posthepatic" ~ "post_liver",
  se$anatomical_origin == "portal_vein" ~ "pre_liver",
  se$anatomical_origin == "right_atrium" ~ "post_liver",
  se$anatomical_origin == "thoracic_inferior_vena_cava" ~ "post_liver",
  TRUE ~ as.character(se$anatomical_origin)
)
se$anatomical_origin_grouped <- factor(se$anatomical_origin_grouped, levels = c("pre_liver", "post_liver"))
```

Subset by feature type. Using only protein coding features
```{r dw-subset-feature-type, eval = TRUE}
use_rows <- rowData(se)$gene_type %in% c('protein_coding')
se <- se[use_rows,]
```

Cell cycle annotation with cyclone, before filtering low abundance genes: *Users should not filter out low-abundance genes before applying cyclone. Even if a gene is not expressed in any cell, it may still be useful for classification if it is phase-specific.*
```{r dw-cc-cyclone}
cc.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", 
    package="scran"))
res <- cyclone(se, cc.pairs)
se$cc_phase_cy <- res$phases
se$cc_g1_score_cy <- res$normalized.scores$G1
se$cc_s_score_cy <- res$normalized.scores$S
se$cc_g2m_score_cy <- res$normalized.scores$G2M
cc_assignments_cy <- res
```


Filtering out low-abundance features. Keeping those features with at least 5 counts in at least 50% of samples of the smallest group.
```{r dw-subset-feature-abundance-hs}
# Not done: it is done at DGE level, for each comparison
use_rows <- filterByExpr(se, group=colData(se)[[params$group]], min.count = 5, min.total.count = 15, min.prop = 0.5)
se <- se[use_rows,]
```

TMM normalization
```{r dw-tmm-hs}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)
```

CPM calculation
```{r dw-cpm-hs}
assay(se, 'cpm') <- edgeR::cpm(assay(se, 'normcounts'))
assay(se, 'logcpm') <- edgeR::cpm(assay(se, 'normcounts'), log = TRUE)
```

Variance stabilizing transformations
```{r dw-vst-hs, eval = FALSE}
assay(se, 'vst') <- vst(assay(se, 'counts'), blind = TRUE)
```


### Additional objects
Selection of HVG
```{r conf-hvg}
# Variance modelling
fit_res <- modelGeneVar(se)
fit_md <- metadata(fit_res)
# Define highly variable genes
hvg <- getTopHVGs(fit_res, prop = 0.1)
hvg_top_n100 <- getTopHVGs(fit_res, n = 100)

# HVG using gene names instead of ensembl IDs
fit_res_ensg <- fit_res
fit_res_ensg$ensg <- rownames(fit_res)
fit_res_ensg <- fit_res_ensg %>% 
  data.frame %>% 
  left_join(rowData(se) %>% data.frame %>% rownames_to_column('ensg') %>% dplyr::select(ensg, gene_name)) %>% 
  filter(!grepl("^ENSG", gene_name)) %>% 
  column_to_rownames('ensg') %>% 
  select(-gene_name)

hvg_top_n100_f <- getTopHVGs(fit_res_ensg, n = 100)

```

Annotation variables and colors
```{r conf-annot}
ann_vars <- strsplit(params$ann_vars, ",") %>% unlist %>% 
  intersect(., colData(se) %>% colnames)
ann_df <- colData(se) %>% data.frame %>% 
  dplyr::select(all_of(ann_vars))

ann_colors <- foreach(i = ann_vars) %do% {
  get_palette(i)
}
names(ann_colors) <- ann_vars
ann_colors <- compact(ann_colors)
```



## Descriptive tables

### By main condition
```{r descriptive-table-main}

use_formula <- as.formula(paste0(ann_vars, "~",  paste(setdiff(ann_vars, params$group), collapse = "+")))
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE)
summary(tab1)
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE, strata=sample_type)
summary(tab1)
```


## PCA
```{r pca-conf}
use_rows <- hvg
mat <- assay(se[use_rows,], 'logcounts')
rownames(mat) <- make.names(rowData(se[use_rows,])$gene_name,unique = T)
# Create annotation, keeping only variables with more than 1 unique value
use_cols <- ann_df %>% dplyr::select(where(~n_distinct(.) > 1)) %>% colnames
p <- PCAtools::pca(mat, metadata = colData(se) %>% data.frame %>% dplyr::select(all_of(use_cols)))
```


### Determine optimum number of PCs to retain
Horn’s parallel analysis and elbow method
```{r pca-elbow, fig.width = 3.1, fig.asp = 0.9}
horn <- parallelPCA(mat)
elbow <- findElbowPoint(p$variance)
# horn$n
# elbow
screeplot(
  p,
  axisLabSize = 8, titleLabSize = 8,
  components = getComponents(p, 1:20),
  vline = c(horn$n, elbow)
) +
  geom_label(aes(x = horn$n + 0.5, y = 25, label = 'Horn\'s', vjust = -1),
             size = geom_text_size) +
  geom_label(aes(x = elbow + 0.5, y = 50, label = 'Elbow method', vjust = -1),
             size = geom_text_size)
```



### Variable correlation
Plot showing the correlation of PCs with sample-level variables. 
```{r pca-eigencor-plot, fig.width = 10.2}
use_rows <- hvg
n_pc <- seq_len(max(c(horn$n, elbow)))
if(length(n_pc) <= 2)
  n_pc <- 1:3

eigencor_res <- eigencorval(p, components = getComponents(p, n_pc), metavars = colnames(p$metadata))

```

```{r pca-eigencor-plot-curated, fig.width = 7.2}
use_rows <- hvg
selected_vars <- c(
  `Anatomical origin` = 'anatomical_origin',
  `CTC type` = 'sample_type',
  `Number of cells` = 'n_cells',
  `Detected genes` = 'detected',
  `Total counts` = 'sum',
  `Mitochondrial %` = 'subsets_Mito_percent',
  `Ribosomal %` = 'subsets_Ribo_percent',
  `G1 score` = 'cc_g1_score_cy',
  `S score` = 'cc_s_score_cy',
  `G2M score` = 'cc_g2m_score_cy'
)

p_curated <- p  
p_curated$metadata <- p_curated$metadata %>% 
  dplyr::rename(all_of(selected_vars))
eigencor_res <- eigencorval(p_curated, components = getComponents(p_curated, n_pc), metavars = names(selected_vars))

# Plot correlation
eigencorplot(p_curated,
             components = getComponents(p_curated, n_pc),
             metavars = names(selected_vars),
             fontCorval = 2,
             cexCorval = 0.8,
             main = 'PC sample variables correlation')


```


## Hierarchical clustering
```{r hclust-conf}
max_pc <- ifelse(ncol(p$rotated) < 10, ncol(p$rotated), 10)
dst <- dist(p$rotated[,1:max_pc], method = 'euclidean')
hc <-  dst %>% hclust(method = 'ward.D2')
dend <- hc %>% as.dendrogram()
column_dend<- dend

## Find Clusters
clusters_original <- cutreeDynamic(hc, distM = as.matrix(dst),
                                   minClusterSize = 5, method="tree")
if(min(clusters_original) == 0) {
  clusters_original <- clusters_original + 1
}

## Add clustering info to se


cluster_info <- data.frame(
  sample_alias = rownames(p$rotated)[order.dendrogram(dend)],
  cluster = as.character(clusters_original[order.dendrogram(dend)]) %>% 
    factor(levels = as.character(unique(clusters_original)))
  )

if(!is.null(se$cluster))
  se$cluster <- NULL
colData(se) <- colData(se) %>% data.frame %>% 
  left_join(cluster_info) %>% 
  mutate(torownames = sample_alias) %>%
  column_to_rownames('torownames') %>% 
  DataFrame

ann_df$cluster <- colData(se)[rownames(ann_df), 'cluster'] %>% as.character
```

### Dendrogram
Plot hierarchical clustering
```{r hclust-plot, fig.width=7.2}
## Sort clusters to the order of the dendrogram:
clusters <- clusters_original[order.dendrogram(dend)]
clusters_numbers <- unique(clusters) - (0 %in% clusters)
n_clusters <- length(clusters_numbers)
cluster_cols <- rainbow_hcl(n_clusters) %>% 
  set_names(1:n_clusters)
column_dend_colored <- dend %>% 
         branches_attr_by_clusters(clusters, values = cluster_cols) 


# Add color lables
bar_var_sorted <- ann_df[[params$group]][order.dendrogram(dend)]
bar_color <- get_palette(params$group)[bar_var_sorted]

# Plot colored dendrogram
plot(column_dend_colored)
colored_bars(colors = bar_color, dend = column_dend_colored, rowLabels = params$group)

# Add cluster color to annotation df
ann_colors$cluster <- cluster_cols
```

### Heatmap 
```{r hclust-heatmap-pca, fig.width=12.5, fig.asp=1}
mat <- p$rotated[,1:max_pc] %>% t
pheatmap(mat, 
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         cluster_cols = TRUE,
         annotation_col = ann_df[colnames(mat),] %>% select('anatomical_origin_grouped'), 
         annotation_colors = ann_colors,
         # fontsize_row = 6,
         show_rownames = TRUE,
         main = 'Heatmap of PC1-10.')
```


### Descriptive table
```{r hclust-descriptive-table}
use_formula <- as.formula(paste0(params$group, " ~ cluster"))
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE)
summary(tab1)
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE, strata=sample_type)
summary(tab1)

use_formula <- as.formula(paste0("cluster", "~", paste(ann_vars, collapse = "+")))
tab1 <- tableby(use_formula, 
                data=colData(se) %>% data.frame,total = FALSE)
summary(tab1)
```


## HVG heatmap 
```{r hvg-exprs-heatmap-hvg-curated, fig.width=4.5, fig.asp=1.8}
use_rows <- hvg_top_n100_f
mat <- assay(se, 'logcounts')[use_rows,]
rownames(mat) <- rowData(se)[use_rows, 'gene_name']
zmat <- t(apply(mat, 1, scale, center = TRUE, scale = TRUE))
colnames(zmat) <- colnames(mat)

# Get sample clustering information
clvar <- 'cluster'
se$uclust <- colData(se)[[clvar]]

# Column annotation
use_annot <- ann_df[colnames(zmat),] %>% 
  select(anatomical_origin_grouped)
use_annot$cluster <- colData(se)[rownames(use_annot), 'uclust'] %>% as.character
heatmap_col_annot_vars <- c('anatomical_origin_grouped')

column_ha_colors <- lapply(heatmap_col_annot_vars, get_palette) %>% 
    set_names(heatmap_col_annot_vars) %>% 
    compact

column_ha <-  HeatmapAnnotation(
    df = use_annot %>% dplyr::select(all_of(heatmap_col_annot_vars)),
    col = column_ha_colors,
    annotation_name_gp = gpar(fontsize = 8),
    simple_anno_size = unit(0.2, "cm"),
    annotation_legend_param = list(
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 8)
    )
  )


ht <- Heatmap(
  mat,
  name = 'Loadings',
  row_names_gp = gpar(fontsize = 6),
  show_row_names = FALSE,
  show_column_names = FALSE,
  show_column_dend = TRUE,
  cluster_rows = TRUE,
  # cluster_columns = hc,
  
  top_annotation = column_ha,
  column_split = use_annot$cluster,
  column_title = "Cluster %s",
  column_title_gp = gpar(fontsize = 6),
  
  
  # row_split = 2,
  
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 6),
    labels_gp = gpar(fontsize = 6),
    ncol = 2
  )
)
draw(
  ht, 
  annotation_legend_side = "bottom"
)

cat("\n\n")

```

## Cluster markers: differental expression
```{r hclust-mk-conf}
clvar <- 'cluster'
se$uclust <- colData(se)[[clvar]]

# Define comparisons
x <- "1"
comp_list <- foreach(x = levels(se$uclust)) %do%{
  y <- setdiff( levels(se$uclust), x)
  list(
    colnames(se)[as.character(se$uclust) == x],
    colnames(se)[as.character(se$uclust)%in% y]
  ) %>% set_names(c(x, paste(y, collapse = "-")))
}
names(comp_list) <- levels(se$uclust)

# RUN edgeR_dge and gse_omnibus
if(params$run_cl_mk){
  # Get ensembl annotation
  ensembl <-  biomaRt::useEnsembl(biomart="genes", dataset="hsapiens_gene_ensembl",
                                  mirror = 'useast')
  gene_desc <- biomaRt::getBM(attributes=c('external_gene_name','description'),
                              filters = 'external_gene_name',
                              values = rowData(se)$gene_name, mart =ensembl) %>%
    dplyr::rename('gene_name' = 'external_gene_name') %>%
    unique
  
  gene_desc <- gene_desc[!duplicated(gene_desc$gene_name),]
  # saveRDS(gene_desc, file = file.path(dirname(params$se_path), 'ensembl_gene_desc.rds'))
  
  # Run differential expression analysis using EdgeR QLF
  use_comp <-  comp_list[[1]]
  dge_list <- foreach(use_comp = comp_list) %do%{
    se_cols <- se$sample_alias %in% unlist(use_comp)
    use_se <- se[,se_cols]
    use_se$group <- ifelse(
      use_se$sample_alias %in% use_comp[[1]],
      names(use_comp)[1],
      names(use_comp)[2]
    )
    
    if(is.null(use_comp$batch)) {
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        design_formula = "~ group",
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
      )
    } else {
      design_formula = paste0("~ ", use_comp$batch," + group")
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        batch_vars = use_comp$batch,
        design_formula = design_formula,
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
      )
    }
    
    # Add gene description
    use_res <- dge$results %>%  left_join(., gene_desc, relationship = "many-to-many")
    dge$results <- use_res %>% 
      filter(!duplicated(feature)) %>% 
      mutate(rownames = feature) %>% 
      column_to_rownames('rownames')
    
    return(dge)
  }
  names(dge_list) <- names(comp_list)
  saveRDS(dge_list, file = file.path(params$output_dir, paste0('dge-',clvar,'.rds')))
}

dge_list <- readRDS(file.path(params$output_dir, paste0('dge-',clvar,'.rds')))
```

<!-- Run template -->
```{r knitr-template-rnaseq-deg-edger}
# Configuration: variable evalauted in differential expression is by default included
heatmap_col_annot_vars <- c('mouse_id', 'sample_type')

# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-deg-edger.Rmd")
  )
```
`r knit(text = src)`

## Cluster markers : ORA
```{r hclust-mk-ora-conf}
# Run ORA gse_omnibus
if(params$run_cl_mk){
  
  # Run functional enrichment
  dge <-  dge_list[[2]]
  gse_list <- foreach(dge = dge_list) %do%{
    gse_res <- gse_omnibus(
      feature_names = dge$results$gene_name,
      p = dge$results$FDR,
      fc = dge$results$logFC,
      fc_thrs = 0.5,
      gmt_files = gmt_files_symbols,
      
      run_enricher = TRUE,
      run_enrichGO = TRUE,
      run_enrichKEGG = FALSE,
      run_enrichReactome = FALSE,
      
      run_GSEA = FALSE,
      run_gseGO = FALSE,
      run_gseKEGG = FALSE,
      run_gseReactome = FALSE,
      
      go_ontologies = c('BP', 'MF'),
      go_simplify_p = 0.05,
      args_gse = list(minGSSize = 10, 
                      maxGSSize = 250, 
                      pvalueCutoff = 1,
                      BPPARAM = BiocParallel::MulticoreParam(1)
      ),
      args_enricher = list(minGSSize = 10, 
                           maxGSSize = 250, 
                           pvalueCutoff = 1, 
                           qvalueCutoff = 1
      )
      
    )
    
    return(gse_res)
    
  }
  names(gse_list) <- names(dge_list)
  saveRDS(gse_list, file = file.path(params$output_dir,paste0('gse_omnibus_ora-',clvar,'.rds')))
}
ora_list <- readRDS(file.path(params$output_dir, paste0('gse_omnibus_ora-',clvar,'.rds')))
``` 

```{r knitr-template-rnaseq-gse-ora}
gse_list <- ora_list
# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-gse-omnibus-ora.Rmd")
  )
```
`r knit(text = src)`

### Dotplot curated
```{r hclust-ora-dotplot-curated, fig.width = 4.2, fig.asp=0.72}
gse_list <- ora_list

x <- rbind(
  gse_list$`2`$enricher$up$msigdb.c2.cp@result %>% mutate(group = 'Cluster 2'),
  gse_list$`1`$enricher$up$msigdb.c2.cp@result %>% mutate(group = 'Cluster 1')
) %>% 
  filter(p.adjust <= 0.05) %>% 
  mutate(
    Description = clean_msigdb_names(Description),
    log10P = -log10(p.adjust),
    Description = fct_reorder(Description, log10P)
  )



x %>% 
  ggplot(aes(x = log10P, y = Description, size = GeneProp, color = group)) +
  geom_point(pch = 19) +
  facet_grid(rows = vars(group), scales = 'free_y', space = 'free_y') +
  scale_y_discrete(labels =  default_labeller(70)) +
  labs(
    y = '',
    x = expression(paste("-lo", g[10],"(Adjusted P value)")),
    size= 'Gene ratio',
    caption = 'Gene sets with adjusted P value ≤ 0.05'
  ) +
  guides(color = 'none') +
  scale_color_manual(values=get_palette('cluster')) +
  scale_size_continuous(range=c(0.4, 2.5)) +
  theme_facet +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
    strip.text.y = element_text(angle = 0, hjust = 0),

    axis.line = element_blank(),
    axis.ticks = element_line(color = "black", linewidth = linewidth_default),
    axis.text.y = element_text(size = 4),
    axis.text.x = element_text(size = 4),

    legend.text = element_text(size = 4),
    legend.title = element_text(size = 4),
    legend.key.width = unit(0.2, "cm")
  )
```


## Integration with Xenograft results
```{r x-HT29_Li.M2-integration-cluster-conf}
# Set the SummarizedExperiment object
use_se <- se
rownames(use_se) <- make.names(rowData(use_se)$gene_name, unique = TRUE)
gsvaAnnotation(use_se) <- SymbolIdentifier("org.Hs.eg.db")

# Selected gene sets from msigdb.c2.reactome
selected_gsets <- list(
  `1` = c('REACTOME_KERATINIZATION', 'REACTOME_DNA_STRAND_ELONGATION', 'REACTOME_DEVELOPMENTAL_CELL_LINEAGES', 'REACTOME_CHEMOKINE_RECEPTORS_BIND_CHEMOKINES', 'REACTOME_INTERFERON_SIGNALING'),
  `2` = c('REACTOME_EUKARYOTIC_TRANSLATION_ELONGATION', 'REACTOME_CYTOSOLIC_TRNA_AMINOACYLATION', 'REACTOME_SLC_MEDIATED_TRANSPORT_OF_AMINO_ACIDS')
)

# Extract gene-sets from GMT file
use_gsets <- readGMT(gmt_files_symbols$msigdb.c2.cp, valueType = 'list')
use_gsets <- use_gsets[unlist(selected_gsets)]

# Generate signatures based on CTC cluster markers. Here we use the results from differential expression between clusters identified by hierarchical clustering. We select only protein coding genes to match data from TCGA. For each group,  gene signature was defined as top-100 genes upregulated in each group (PValue < 0.05 and logFC > 0.5; genes ranked by sign(logFC) × -log10(PValue)). 

topn <- 100
p_thresh <- 0.05
logfc_thresh <- 0.5
gene_sets <- list()

dge_list <- readRDS('./output/rnaseq/HT29_Li.M2/exploratory/dge-cluster_rmpc1.rds')
use_dge <-  dge_list[[1]]
gene_sets <- foreach(use_dge = dge_list) %do% {
  use_dge$results |>
    filter(logFC >= logfc_thresh & PValue < p_thresh) |> 
    filter(gene_type == 'protein_coding') |> 
    mutate(rank = sign(logfc_thresh) * -log10(PValue)) |>
    arrange(desc(rank)) |> 
    head(topn) |>
    pull(gene_name)
}
names(gene_sets) <- paste0('Mouse cluster',names(dge_list))

use_gsets <- c(use_gsets, gene_sets)

# Run GSVA
gsva_params <- gsvaParam(
  assay(use_se, "logcpm"), 
  use_gsets
  )
gsva_res <- gsva(gsva_params)

# Extract GSVA scores and add to colData
gsva_df <- gsva_res %>% t() %>%  data.frame %>% 
    rownames_to_column('sample_alias') %>%
  mutate(torow = sample_alias) %>% column_to_rownames('torow')
colData(use_se) %<>% 
  data.frame %>% 
  rownames_to_column('torow') %>%
  left_join(gsva_df, by = 'sample_alias') %>% 
  column_to_rownames('torow') %>%
  DataFrame
```



### Whisker plots
```{r x-HT29_Li.M2-integration-whisker-cluster, fig.width=2.1, fig.asp=1.8}
xvar <- 'uclust'

use_data <- colData(use_se) %>% 
  data.frame %>% 
  pivot_longer(cols = starts_with(c("REACTOME", "module")), names_to = 'gset', values_to = 'gsva_score') %>% 
  mutate(
    gset = clean_msigdb_names(gset),
    gset = gsub("^(.{10}[^ ]*)\\s", "\\1\n",gset)
  )

gset_test <- use_data %>% 
  group_by(gset) %>% 
  wilcox_test(data = ., formula = gsva_score ~ anatomical_origin_grouped) %>% 
  adjust_pvalue(method = "BH") %>% 
  arrange(p)



my_comparisons <- combn(as.character(unique(use_data[[xvar]])), 2, simplify = FALSE)
use_data %>%
  mutate(gset = factor(gset, levels = gset_test$gset)) %>%
  ggplot(aes(x = .data[[xvar]], y = gsva_score)) + 
  geom_boxplot( aes(fill = .data[[xvar]]), outliers = FALSE, alpha = 0.7, width = 0.5, 
                linewidth = linewidth_default,  median.linewidth = linewidth_default,
                color = 'grey30') + 
  # geom_hline(yintercept = 0, linetype = 'dashed', linewidth = linewidth_default) +
  facet_wrap(~gset, ncol = 2, scales = 'free_y') +
  # geom_jitter(width = 0.2, aes(color = .data[[params$group]])) +
  geom_jitter(width = 0.2, alpha = 0.8, size = 0.5) +
  # scale_color_manual(values = get_palette('anatomical_origin_grouped')) +
  scale_fill_manual(values = get_palette(xvar)) +
  labs(
    x = '',
    y = 'GSVA score'
  ) +
  guides(color = 'none', fill = 'none') +
  theme_facet +
  theme(
    strip.text = element_text(size = 2.5)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
  stat_compare_means(comparisons = my_comparisons, size = use_font_size * 0.352777778, bracket.size  = linewidth_default)

```



### Heatmap
```{r x-HT29_Li.M2-integration-heatmap-cluster, fig.width=7.1}
mat <- gsva_res
rownames(mat) <- clean_msigdb_names(rownames(mat))

# Column annotation
column_ha <-  HeatmapAnnotation(
    df = colData(use_se)[colnames(mat),] %>% data.frame %>% select(all_of(c('uclust'))),
    col = list(
      uclust = get_palette('uclust')
    )
  )

htz <- Heatmap(
  mat,
  name= 'GSVA\nscore',
  
  row_names_side = 'left',
  # show_row_dend = FALSE,
  row_dend_side = "right",
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  
  top_annotation = column_ha,
  )

draw(
    htz, 
    annotation_legend_side = "bottom"
  )

```



