---
title:  Analysis of xenograft samples
subtitle: Differential gene expression
author: Francesc Castro-Giner
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  se_files: ./data/rnaseq/HT29_Li.M2/se_hs.rds,./data/rnaseq/HT29/se_hs.rds
  output_dir: ./output/rnaseq/xenograft/hs/differential_expression_edger/donor
  group: condition
  min_counts: 5
  min_present_prop: 0.70
  msigdb_dir: ./data/resources/MSigDB
  msigdb_vs: v2026.1.Hs
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(circlize)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
source('./code/R-functions/pairwise_terms_jc.r')
source('./code/R-functions/eigencorval.R')
```

Load MSigDB gene sets
```{r load-msigdb}
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/c2.cp.', params$msigdb_vs, '.symbols.gmt')
)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Process param variables
```{r process-params}
se_files <- strsplit(params$se_files, ",") %>% unlist
```

## Configure analyses

### Data wrangling
Load SE object and remove outliers
```{r dw-load-se}
# Load files
se_list <- foreach(i=se_files) %do% {
  ise <- readRDS(i)
  rowData(ise)$mean <- NULL
  rowData(ise)$detected <- NULL
  ise$collection_time <- NULL
  ise$mouse_id <- as.character(ise$mouse_id)
  ise$collection_date <- NULL
  ise$ctc_cluster_size <- as.numeric(ise$ctc_cluster_size)
  ise$outlier_high_subsets_Mito_percent <- as.logical(ise$outlier_high_subsets_Mito_percent)
  ise$outlier_low_lib_size <- as.logical(ise$outlier_low_lib_size)
  ise$outlier_low_n_features <- as.logical(ise$outlier_low_n_features)
  ise
}

# Define row intersection
use_genes <- foreach(i=se_list) %do% {
  rownames(i)
} %>% unlist
use_genes <- table(use_genes) %>% data.frame %>% 
  filter(Freq == length(se_files)) %>% pull(use_genes) %>% as.character

# Combine SE
se <- se_list[[1]][use_genes,]
if(length(se_list) > 1) {
  for (i in 2:length(se_files)) {
    ise <- se_list[[i]]
    new_colData <- bind_rows(colData(se) %>% data.frame, 
                             colData(ise) %>% data.frame) %>% 
      DataFrame
    new_counts <- cbind(assay(se, 'counts'), assay(ise[rownames(se),], 'counts'))
    se <- SummarizedExperiment(
      assays = SimpleList(counts = new_counts),
      colData = new_colData,
      rowData = rowData(se)
    )
  }
}

# Update metrics
se <- addPerFeatureQCMetrics(
  se,
  threshold = 3
  )

colData(se) %<>% data.frame %>% 
  mutate(
    total_features_o = nrow(se),
    pct_detected_o = 100*detected/total_features_o
  ) %>% DataFrame


# Rename anatomical origin
se$anatomical_origin_grouped <- case_when(
  se$anatomical_origin == "heart_puncture" ~ "post_liver",
  se$anatomical_origin == "inferior_vena_cava_posthepatic" ~ "post_liver",
  se$anatomical_origin == "portal_vein" ~ "pre_liver",
  se$anatomical_origin == "right_atrium" ~ "post_liver",
  se$anatomical_origin == "thoracic_inferior_vena_cava" ~ "post_liver",
  TRUE ~ as.character(se$anatomical_origin)
)
```


Subset by feature type. Using only protein coding features
```{r dw-subset-feature-type, eval = TRUE}
use_rows <- rowData(se)$gene_type %in% c('protein_coding')
se <- se[use_rows,]
```

Filtering out low-abundance features. Keeping those features with at least 5 counts in at least 50% of samples of the smallest group.
```{r dw-subset-feature-abundance, eval = FALSE}
# Not done: it is done at DGE level, for each comparison
use_rows <- filterByExpr(se, group=colData(se)[[params$group]], min.count = 5, min.total.count = 15, min.prop = 0.5)
se <- se[use_rows,]
```

TMM normalization
```{r dw-tmm}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)
```

CPM calculation
```{r dw-cpm}
assay(se, 'cpm') <- edgeR::cpm(assay(se, 'normcounts'))
assay(se, 'logcpm') <- edgeR::cpm(assay(se, 'normcounts'), log = TRUE)
```

Variance stabilizing transformations
```{r dw-vst, eval = FALSE}
assay(se, 'vst') <- vst(assay(se, 'counts'), blind = TRUE)
```



### Define of comparisons

#### Process data
Define initial comparisons without batch correction
```{r conf-comparisons-unbathced}
x <- colData(se) %>% data.frame
comp_list <- list(
  `29m2-o-29-pv-ctc_cluster` = list(
    `HT29LiM2` = x %>% 
      filter(donor == 'HT29_Li.M2' & anatomical_origin == 'portal_vein' & sample_type == 'ctc_cluster') %>% 
      pull(sample_alias),
    `H29` = x %>% 
      filter(donor == 'HT29' & anatomical_origin == 'portal_vein' & sample_type == 'ctc_cluster') %>% 
      pull(sample_alias)
  )
)

comp_list_description <- data.frame(
  comparison = names(comp_list),
  comparison_long = c(
    "HT29LiM2-over-HT29-PV-CTC_clusters"
  )
)

```

Add comparisons with batch correction. Only using batch levels with samples in both categories and comparisons with more than 2 levels of the batch variable
```{r conf-comparisons-bathced}
batch_var <- 'donor'
batch_levels <- unique(x[,batch_var])
for(cn in names(comp_list)) {
  i <- comp_list[[cn]]
  sample_batch_levels <- x %>% 
    filter(sample_alias %in% i[[1]]) %>% 
    pull(batch_var) %>% 
    unique
  ctrl_batch_levels <- x %>% 
    filter(sample_alias %in% i[[2]]) %>% 
    pull(batch_var) %>% 
    unique
  use_batch_levels <- intersect(sample_batch_levels, ctrl_batch_levels)
  sample_samples <-  x %>% 
    filter(sample_alias %in% i[[1]]) %>% 
    filter(.[[batch_var]] %in% use_batch_levels) %>% 
    pull(sample_alias)
  ctrl_samples <-  x %>% 
    filter(sample_alias %in% i[[2]]) %>% 
    filter(.[[batch_var]] %in% use_batch_levels) %>% 
    pull(sample_alias)
  if(length(use_batch_levels) > 1 & 
      length(sample_samples) > 0 & 
      length(ctrl_samples) > 0) {
    # Add comparison to list of comparison
    res <- list(
      sample = sample_samples,
      ctrl = ctrl_samples,
      batch = batch_var
    )
    names(res)[1:2] <- names(i)[1:2]
    comp_list[[paste0(cn, '-', batch_var)]] <- res
    # Add comparison long name description
    new_desc <- comp_list_description %>% 
      filter(comparison == cn) %>%
      mutate(comparison_long = paste(comparison_long, 'batch', batch_var),
             comparison = paste0(cn, '-', batch_var))
    comp_list_description <- rbind(comp_list_description, new_desc)
  }  
}
```


#### Table of comparisons

We filter comparisons with n < 3 samples in each group
```{r table-comparisons}
i = names(comp_list)[1]
comp_df <- foreach(i = names(comp_list), .combine = rbind) %do% {
  icomp <- comp_list[[i]]
  data.frame(
    comparison = i,
    `Sample Group (n)` =  length(icomp[[1]]),
    `Reference Group (n)` = length(icomp[[2]]),
    check.names = FALSE
  )
}

# filter comparisons were we have at least 3 samples in each group
comp_df %<>% 
  mutate(Evaluated = `Sample Group (n)` >= 3 & `Reference Group (n)` >= 3)
kepp_comp <- comp_df %>% filter(Evaluated) %>% pull(comparison)
comp_list <- comp_list[kepp_comp]

# Add long names
comp_df <- comp_df %>% left_join(comp_list_description) %>% 
  mutate(comparison = comparison_long) %>%
  dplyr::select(-comparison_long)

comp_df %>%
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'List of comparisons for differential expression. Comparisons with n < 3 samples in each group won\'t be evaluated',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            ))

```

## Data generation

### Run differential expression analysis
Get Ensembl gene info
```{r de-ensembl, eval = FALSE}
ensembl <-  biomaRt::useEnsembl(biomart="genes", dataset="hsapiens_gene_ensembl",
                                mirror = 'useast')
gene_desc <- biomaRt::getBM(attributes=c('external_gene_name','description'),
                              filters = 'external_gene_name',
                              values = rowData(se)$gene_name, mart =ensembl) %>%
    dplyr::rename('gene_name' = 'external_gene_name') %>%
    unique

gene_desc <- gene_desc[!duplicated(gene_desc$gene_name),]
saveRDS(gene_desc, file = file.path(dirname(params$output_dir), 'ensembl_gene_desc.rds'))
```

```{r de-ensembl-load, eval = FALSE}
gene_desc <- readRDS(file = file.path(dirname(params$output_dir), 'ensembl_gene_desc.rds'))
```

Run differential expression analysis using EdgeR QLF
```{r edgeR-QLF-run, eval = FALSE}
use_comp <-  comp_list[[1]]
dge_list <- foreach(use_comp = comp_list) %do%{
  se_cols <- se$sample_alias %in% unlist(use_comp)
  use_se <- se[,se_cols]
  use_se$group <- ifelse(
    use_se$sample_alias %in% use_comp[[1]],
    names(use_comp)[1],
    names(use_comp)[2]
  )
  
  if(is.null(use_comp$batch)) {
    dge <- edgeR_dge(
      use_se,
      # Design configuration for differential expression
      group_var =  'group',
      group_sample = names(use_comp)[1],
      group_ref = names(use_comp)[2],
      design_formula = "~ group",
      coef = 'last',
      # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
      assay_to_DGEList = 'counts',
      assay_to_row_filter = "counts",
      # Feature filtering parameters
      use_filterByExpr = TRUE,
      min_counts = params$min_counts,
      min_present_prop = params$min_present_prop,
      # EdgeR workflow configuration
      run_calcNormFactors = 'TMM',
      estimateDisp_robust = TRUE,
      estimateDisp_trend.method = "locfit",
      glm_approach = "QLF",
      # Output configuration
      adjust_method = 'BH'
      )
    } else {
      design_formula = paste0("~ ", use_comp$batch," + group")
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        batch_vars = use_comp$batch,
        design_formula = design_formula,
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
        )
    }
  
  # Add gene description
  use_res <- dge$results %>%  left_join(., gene_desc, relationship = "many-to-many")
  dge$results <- use_res %>% 
    filter(!duplicated(feature)) %>% 
    mutate(rownames = feature) %>% 
    column_to_rownames('rownames')
  
   return(dge)
}

names(dge_list) <- names(comp_list)

# Save objects
saveRDS(dge_list, file = file.path(params$output_dir, 'dge-edgeR_QLF.rds'))

###########
#   use_samples = NULL
#   # Desing configuration for differential expression
#   group_var = NULL
#   group_sample = NULL
#   group_ref = NULL
#   numeric_covar = NULL
#   batch_vars = NULL
#   design_formula = NULL
#   contrast = NULL
#   coef = NULL
#   # Conversion from SingleCellExpeiment to DGEList
#   spike_normalization = FALSE
#   assay_to_DGEList = 'counts'
#   assay_to_row_filter = "tpm"
#   use_colData = NULL
#   use_rowData = NULL
#   # Feature filtering parameters
#   use_filterByExpr = TRUE
#   min_counts = 1
#   min_present_prop = 0.40
#   # EdgeR workflow configuration
#   run_calcNormFactors = 'TMM'
#   estimateDisp_robust = FALSE
#   estimateDisp_trend.method = "locfit"
#   glmQLFit_robust = FALSE
#   glm_approach = "QLF"
#   # Output configuration
#   adjust_method = 'BH'
#   assays_from_SingleCellExperiment = NULL
# # 
#     object = use_se
#     # Desing configuration for differential expression
#     group_var =  'group'
#     group_sample = names(use_comp)[1]
#     group_ref = names(use_comp)[2]
#     batch_vars = 'donor'
#     design_formula = "~ donor + group"
#     coef = 'last'
#     # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
#     assay_to_DGEList = 'counts'
#     assay_to_row_filter = "counts"
#     # Feature filtering parameters
#     use_filterByExpr = TRUE
#     min_counts = params$min_counts
#     min_present_prop = params$min_present_prop
#     # EdgeR workflow configuration
#     run_calcNormFactors = 'TMM'
#     estimateDisp_robust = TRUE
#     estimateDisp_trend.method = "locfit"
#     glm_approach = "QLF"
#     # Output configuration
#     adjust_method = 'BH'

```

### Run functional enrichment
```{r gse-run, eval = FALSE}
dge_list <- readRDS(file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
dge <-  dge_list[[2]]

gse_list <- foreach(dge = dge_list) %do%{
  gse_res <- gse_omnibus(
    feature_names = dge$results$gene_name,
    p = dge$results$FDR,
    fc = dge$results$logFC,
    fc_thrs = 0.5,
    gmt_files = gmt_files_symbols,
    
    run_enricher = TRUE,
    run_enrichGO = TRUE,
    run_enrichKEGG = FALSE,
    run_enrichReactome = FALSE,
    
    run_GSEA = TRUE,
    run_gseGO = TRUE,
    run_gseKEGG = FALSE,
    run_gseReactome = FALSE,
    
    go_ontologies = c('BP', 'MF'),
    go_simplify_p = 0.05,
    args_gse = list(minGSSize = 10, 
                    maxGSSize = 250, 
                    pvalueCutoff = 1,
                    BPPARAM = BiocParallel::MulticoreParam(1)
    ),
    args_enricher = list(minGSSize = 10, 
                         maxGSSize = 250, 
                         pvalueCutoff = 1, 
                         qvalueCutoff = 1
    )
    
    )
  
   return(gse_res)

}
names(gse_list) <- names(dge_list)

# Save objects
saveRDS(gse_list, file = file.path(params$output_dir,'gse_omnibus-edgeR_QLF.rds'))
```



### Load data
```{r dge-gse-load}
dge_list <- readRDS(file = file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
gse_list <- readRDS(file = file.path(params$output_dir, 'gse_omnibus-edgeR_QLF.rds'))
```



## Differential Gene Expression
<!-- Run template -->
```{r knitr-template-rnaseq-deg-edger}
# Configuration: variable evalauted in differential expression is by default included
heatmap_col_annot_vars <- c('donor', 'anatomical_origin')

# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-deg-edger.Rmd")
  )
```
`r knit(text = src)`


## Functional enrichment : GSEA
```{r gse-conf}
gse_collections <- list(
  GSEA = names(gmt_files_symbols)
)
```


### Download results
<!-- Write GSE results to xlsx file -->
```{r write-gse_list-xlsx}
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'

file_xlsx_list <- list()
for(sr in names(gse_collections)) {
  file_xlsx_list[[sr]] <- list()
  for(gc in gse_collections[[sr]]) {
    # Generate file
    file_xlsx <- file.path('./docs/file', 
                           rmd_file, 
                           paste0('gse_', tolower(sr), '_', tolower(gc), '.xlsx')
                           )
    file_xlsx_list[[sr]][[gc]] <- file_xlsx
    dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)
    
    # Generate workbook
    wb <- createWorkbook()
    for(i in names(gse_list)) {
      addWorksheet(wb, i)
      if(gc != sr){
        if(!is.null(gse_list[[i]][[sr]][[gc]])) {
          res <- gse_list[[i]][[sr]][[gc]]@result
          writeData(wb, i, res)
        }
      }
      if(gc == sr){
        if(!is.null(gse_list[[i]][[sr]])) {
          res <- gse_list[[i]][[sr]]@result
          writeData(wb, i, res)
        }
      }
    }
    saveWorkbook(wb, file_xlsx, TRUE)
  }
}
```

<!-- Change gse_list names after writing to xlsx -->
```{r change-gse-names}
if(exists('comp_list_description')) {
  comp_names <- comp_list_description$comparison_long %>% 
    set_names(comp_list_description$comparison)
  
  comp_list_backup <- comp_list
  names_gse_list_backup <- names(gse_list)
  
  names(comp_list) <- comp_names[names(comp_list)]
  names(gse_list) <- comp_names[names(gse_list)]
}
```

<!-- Write links to download GSE results -->
```{r write-links-gse-2, echo = FALSE}
for(sr in names(file_xlsx_list)) {
  for(gc in names(file_xlsx_list[[sr]])) {
    file_xlsx <- gsub("docs/", "", file_xlsx_list[[sr]][[gc]])
    if(sr == 'GSEA') {
      cat(paste0("- [**Gene set enrichment analysis: ", gc, "**](", file_xlsx, ")"), '\n\n')
    }
  }
}
```






### Similarity Matrix
<!-- Similarity matrix adn clsutering -->
```{r 29m2-o-29-pv-ctc_cluster}
comp <- "HT29LiM2-over-HT29-PV-CTC_clusters"
p_thrs <- 0.01
gse_collections <- list(
  GSEA = "msigdb.c2.cp"
)

# Generate JC similarity matrix
jc_matrix_list <- foreach(use_comp = comp) %do% {
  data_gse <- foreach(sr = 'GSEA') %do% {
    data_by_collection <- foreach(gc = gse_collections[[sr]], .combine = rbind)  %do% {
      if(sr == 'GSEA') {
        x <- gse_list[[use_comp]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          arrange(p.adjust)
      }
      return(x)
    }
    return(data_by_collection)
  }

  data_gse <- purrr::compact(data_gse)
  gse_df <- bind_rows(data_gse)
  if(nrow(gse_df) > 1) {
    gse_df <- bind_rows(data_gse) %>% 
      mutate(
        Description = clean_msigdb_names(Description),
        Description = ifelse(grepl("^GO:", ID), paste0("GO:", Description), Description)
      )

    
    # Data frame to plot
    res_top <- gse_df %>%
      mutate(
        log.p.adjust = -log10(p.adjust),
        Description = factor(Description)
      )
    
    # Calculate jaccard similarity matrix
    gset_core_enrichment <- foreach(i = levels(res_top$Description)) %do% {
      res_top %>% filter(Description == i) %>% pull(core_enrichment) %>%
        paste(collapse = '/') %>%
        strsplit('/') %>% unlist %>% unique
    }
    names(gset_core_enrichment) <- levels(res_top$Description)
    
    use_terms <- res_top$Description %>% unique
    jc_matrix <- pairwise_terms_jc(gset_core_enrichment[use_terms])
  } else {
    jc_matrix <- NULL
  }  
  return(jc_matrix)
}

names(jc_matrix_list) <- comp
jc_matrix_list <- purrr::compact(jc_matrix_list)



# Generate Clsutering and Dendrogram from JC similarity matrix
library(dynamicTreeCut)
use_mat <- jc_matrix_list[[comp]]
dst <- dist(use_mat, method = 'euclidean')
hc <-  dst %>% hclust(method = 'ward.D2')
dend <- hc %>% as.dendrogram()
column_dend <- dend
hc_sample_order <- cbind(Description = hc$labels, hc_order = hc$order) %>% 
  data.frame %>% 
  mutate(hc_order = as.numeric(hc_order))
## Find Clusters
clusters_original <- cutreeDynamic(hc, 
                                   distM = as.matrix(dst), method="hybrid", 
                                   minClusterSize = 3)
## Combine clustering info
cluster_info <- data.frame(
  Description = rownames(use_mat)[order.dendrogram(dend)],
  cluster = as.character(clusters_original[order.dendrogram(dend)]) %>% 
    factor(levels = as.character(unique(clusters_original)))
  ) %>% 
  left_join(., hc_sample_order, by = 'Description')
```


```{r 29m2-o-29-pv-ctc_cluster-jc-heatmap, fig.width=6.2, fig.height=5.5}
library(NbClust)
use_mat <- jc_matrix_list[[comp]]
colnames(use_mat) <- colnames(use_mat) %>% clean_msigdb_names %>% ep_str_wrap(70)
rownames(use_mat) <- rownames(use_mat) %>% clean_msigdb_names  %>% ep_str_wrap(70)

# Define number of clusters
use_max.nc <- ifelse(nrow(use_mat) <= 15, 5, round(nrow(use_mat) * 0.1))
nbclust_res <- NbClust(data = NULL, diss = as.dist(1-use_mat), distance = NULL, 
                       min.nc = 2, max.nc = use_max.nc, method = "ward.D",
                       index = 'silhouette')


n_split <- nbclust_res$Best.nc['Number_clusters']


col_fun <- colorRamp2(seq(0, 1, length.out = 4), brewer.pal(4, "GnBu"))


ht <- Heatmap(
  use_mat,
  name = 'Jaccard index',
  column_split = n_split,
  row_split = n_split,
  column_title = NULL,
  row_title = NULL,
  row_names_gp = gpar(fontsize = 4),
  col = col_fun,
  show_column_dend = FALSE,
  show_column_names = FALSE,
  border = TRUE,
  # top_annotation = ha_top,
  heatmap_legend_param = list(title_gp = gpar(fontsize = 6),
                              labels_gp = gpar(fontsize = 6)
  ),
  width = unit(3, "cm"),
  
  clustering_method_rows = "ward.D",
  clustering_method_columns = "ward.D"
) 

draw(ht)


ht_clusters_list <- foreach(i = row_order(ht)) %do% {
  rownames(use_mat)[i]
}
names(ht_clusters_list) <- 1:length(ht_clusters_list)

ht_clusters <- data.frame(
  cluster = rep(names(ht_clusters_list), lengths(ht_clusters_list)),
  Description = unlist(ht_clusters_list, use.names = FALSE)
)

```


### Dotplot
```{r curated-29m2-o-29-pv-ctc_cluster-gsea-dotplot}
# Process groups
group_levels_ord <- c(
  "Translation & stress response",
  "Coagulation & Complement cascade",
  "Cytoskeleton, cell-cell adhesion",
  "DNA damage repair & G2/M Checkpoints",
  "G1/S transition & DNA synthesis",
  "Cell cycle & RNA splicing",
  "Cholesterol metabolism",
  "Chromatin remodelling",
  "DNA replication",
  "Nuclear Export & mRNA transport",
  "rRNA transcription",
  "Histone modification",
  "Mitotic spindle & chromatid segregation",
  "Other"
) %>% rev

cluster_info <- nbclust_res$Best.partition %>% data.frame %>%
  rownames_to_column('Description') %>%
  dplyr::rename(cluster = '.') %>%
  arrange(cluster)


grouped_pathways <- gse_list[[comp]]$GSEA$msigdb.c2.cp@result %>%
  filter(p.adjust <= p_thrs) %>%
   mutate(
    Description = clean_msigdb_names(Description) %>% ep_str_wrap(70)
  ) %>%
  left_join(cluster_info) %>%
  mutate(
    cluster_sign = paste(cluster, .sign, sep = '_')
  ) %>% 
  dplyr::select(Description, cluster, cluster_sign) %>% 
  mutate(
    group = case_when(
      cluster_sign == "10_activated" ~ "Translation & stress response",
      cluster_sign == "9_activated" ~ "Coagulation & Complement cascade",
      cluster_sign == "1_activated" ~ "Cytoskeleton, cell-cell adhesion",
      cluster_sign == "4_supressed" ~ "DNA damage repair & G2/M Checkpoints",
      cluster_sign == "1_supressed" ~ "Cell cycle & RNA splicing",
      cluster_sign == "3_supressed" ~ "G1/S transition & DNA synthesis",
      cluster_sign == "7_supressed" ~ "Cholesterol metabolism",
      cluster_sign == "6_supressed" ~ "Chromatin remodelling",
      cluster_sign == "12_supressed" ~ "DNA replication",
      cluster_sign == "5_supressed" ~ "Nuclear Export & mRNA transport",
      cluster_sign == "8_supressed" ~ "rRNA transcription",
      cluster_sign == "11_supressed" ~ "Histone modification",
      cluster_sign == "2_supressed" ~ "Mitotic spindle & chromatid segregation",
      TRUE ~ 'Other'
    ),
    group_rev = factor(group, levels = rev(group_levels_ord)),
    group = factor(group, levels = group_levels_ord)
  )



x <- gse_list[[comp]]$GSEA$msigdb.c2.cp@result %>% 
  filter(p.adjust <= p_thrs) %>%
  mutate(
    log10P = -log10(p.adjust),
    Description = clean_msigdb_names(Description)
    ) %>% 
  left_join(grouped_pathways, by = 'Description')  %>% 
  filter(!is.na(cluster)) %>%
  mutate(
    Description = ep_str_wrap(Description, 70),
    # Description = fct_reorder(Description, hc_order),
    Description = fct_reorder(Description, NES)
    # cluster = fct_reorder(Description, NES)
  )
xlim <- c(-max(abs(x$NES)), max(abs(x$NES)))
```


```{r curated-29m2-o-29-pv-ctc_cluster-gsea-dotplot-named, fig.width = 7.2, fig.asp=1.2}
x %>% 
  ggplot(aes(x = NES, y = Description, size = log10P, color = NES)) +
  geom_point(pch = 19) +
  facet_grid(rows = vars(group_rev), scales = 'free_y', space = 'free_y') +
  scale_y_discrete(labels =  default_labeller(60)) +
  geom_vline(xintercept = 0, lty = 3) +
  labs(
    y = '',
    size = expression(paste("-lo", g[10],"(Adjusted P value)")),
    caption = 'Gene sets with adjusted P value ≤ 0.05'
  ) +
  scale_size_continuous(range=c(0.8, 2)) +
  scale_color_gradient2(trans = "reverse") +
  scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                     limits = xlim) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
    strip.text.y = element_text(angle = 0, hjust = 0),
    
    axis.line = element_blank(),
    axis.ticks = element_line(color = "black", linewidth = linewidth_default),
    axis.text.y = element_text(size = 4),
    axis.text.x = element_text(size = 4),
    
    legend.text = element_text(size = 4), 
    legend.title = element_text(size = 4),
    legend.key.width = unit(0.2, "cm")
  )
```

### Dotplot simplified

```{r curated-29m2-o-29-pv-ctc_cluster-gsea-dotplot-simplified, fig.width = 2.2, fig.asp=0.68}
x %>% 
  group_by(group, .sign) %>% 
  summarise(NES = mean(NES), log10P = mean(log10P)) %>% 
  ungroup() %>%
  ggplot(aes(x = NES, y = group, size = log10P, color = NES)) +
  facet_grid(rows = vars(.sign), scales = 'free_y', space = 'free_y') +
  geom_point(pch = 19) +
  scale_y_discrete(labels =  default_labeller(80)) +
  geom_vline(xintercept = 0, lty = 3) +
  labs(
    x = 'Average\nNES',
    y = '',
    color = 'Average NES',
    size = expression(paste("Mean\n", "-lo", g[10],"(Adjusted P value)"))
  ) +
  scale_size_continuous(range=c(0.8, 2)) +
  scale_color_gradient2(trans = "reverse") +
  scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                     limits = xlim) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
    # strip.text.y = element_text(angle = 0, hjust = 0),
    strip.text.y = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_line(color = "black", linewidth = linewidth_default),
    axis.text.y = element_text(size = 4),
    axis.text.x = element_text(size = 4),
    
    legend.text = element_text(size = 4), 
    legend.title = element_text(size = 4),
    legend.key.width = unit(0.2, "cm")
  )

```

