---
title: Analysis of CTCs from HT29 and HT29.Li_M2 model
subtitle: Explorative analysis of human reads
author: Francesc Castro-Giner
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  se_path: ./data/rnaseq/HT29_Li.M2/se_hs.rds,./data/rnaseq/HT29/se_hs.rds
  group: anatomical_origin
  min_counts: 5
  min_present_prop: 0.70
  msigdb_dir: ./data/resources/MSigDB
  msigdb_vs: v2026.1.Hs
  ann_vars: anatomical_origin,sample_type,n_cells,detected,sum,subsets_Mito_percent,subsets_Ribo_percent,mouse_id,cc_g1_score_cy,cc_s_score_cy,cc_g2m_score_cy,donor
  run_cl_mk: FALSE
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"), fig.width = 3.5, fig.asp = 0.618)

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)
library(arsenal)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(circlize)
library(GSVA)
library(babelgene)
library(dynamicTreeCut)
library(dendextend)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
source('./code/R-functions/pairwise_terms_jc.r')
source('./code/R-functions/eigencorval.R')
clean_msigdb_names <- function(x) x %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Define additional functions
```{r add-functions}
clean_sample_alias <- function(x) 
  gsub('o[0-9]*_[0-9]*-|_gDNA.*', '', x) %>% gsub(".*-MS", "MS", .) %>% gsub("_S[0-9]+_R[0-9].*", "", .) %>% gsub('__STARpass1', '', .) %>% gsub(".k2report", "", .) %>% gsub(".*-|.disambiguated.*", "", .)
```

Load MSigDB gene sets
```{r load-msigdb}
gmt_files_symbols <- list(
  msigdb.h = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/h.all.', params$msigdb_vs, '.symbols.gmt'),
  msigdb.c2.cp = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/c2.cp.', params$msigdb_vs, '.symbols.gmt')
)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Process param variables
```{r process-params}
se_files <- strsplit(params$se_path, ",") %>% unlist
```

## Data wrangling

Load SE object and remove outliers
```{r dw-load-se}
# Load files
se_list <- foreach(i=se_files) %do% {
  ise <- readRDS(i)
  rowData(ise)$mean <- NULL
  rowData(ise)$detected <- NULL
  ise$collection_time <- NULL
  ise$mouse_id <- as.character(ise$mouse_id)
  # ise$order_id_b_fabric <- as.character(ise$order_id_b_fabric)
  # ise$ctc_cluster_id <- as.character(ise$ctc_cluster_id)
  ise$collection_date <- NULL
  ise$ctc_cluster_size <- as.numeric(ise$ctc_cluster_size)
  ise$outlier_high_subsets_Mito_percent <- as.logical(ise$outlier_high_subsets_Mito_percent)
  ise$outlier_low_lib_size <- as.logical(ise$outlier_low_lib_size)
  ise$outlier_low_n_features <- as.logical(ise$outlier_low_n_features)
  ise
}

# Define row intersection
use_genes <- foreach(i=se_list) %do% {
  rownames(i)
} %>% unlist
use_genes <- table(use_genes) %>% data.frame %>% 
  filter(Freq == length(se_files)) %>% pull(use_genes) %>% as.character

# Combine SE
se <- se_list[[1]][use_genes,]
if(length(se_list) > 1) {
  for (i in 2:length(se_files)) {
    ise <- se_list[[i]]
    new_colData <- bind_rows(colData(se) %>% data.frame, 
                             colData(ise) %>% data.frame) %>% 
      DataFrame
    new_counts <- cbind(assay(se, 'counts'), assay(ise[rownames(se),], 'counts'))
    se <- SummarizedExperiment(
      assays = SimpleList(counts = new_counts),
      colData = new_colData,
      rowData = rowData(se)
    )
  }
}

# Update metrics
se <- addPerFeatureQCMetrics(
  se,
  threshold = 3
  )

colData(se) %<>% data.frame %>% 
  mutate(
    total_features_o = nrow(se),
    pct_detected_o = 100*detected/total_features_o
  ) %>% DataFrame


# Rename anatomical origin
se$anatomical_origin_grouped <- case_when(
  se$anatomical_origin == "heart_puncture" ~ "post_liver",
  se$anatomical_origin == "inferior_vena_cava_posthepatic" ~ "post_liver",
  se$anatomical_origin == "portal_vein" ~ "pre_liver",
  se$anatomical_origin == "right_atrium" ~ "post_liver",
  se$anatomical_origin == "thoracic_inferior_vena_cava" ~ "post_liver",
  TRUE ~ as.character(se$anatomical_origin)
)
```

Cell cycle annotation with cyclone, before filtering low abundance genes: *Users should not filter out low-abundance genes before applying cyclone. Even if a gene is not expressed in any cell, it may still be useful for classification if it is phase-specific.*
```{r dw-cc-cyclone}
cc.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", 
    package="scran"))
res <- cyclone(se, cc.pairs)
se$cc_phase_cy <- res$phases
se$cc_g1_score_cy <- res$normalized.scores$G1
se$cc_s_score_cy <- res$normalized.scores$S
se$cc_g2m_score_cy <- res$normalized.scores$G2M
cc_assignments_cy <- res
```

Filtering out low-abundance features. Keeping those features with at least 5 counts in at least 50% of samples of the smallest group.
```{r dw-subset-feature-abundance-hs}
# Not done: it is done at DGE level, for each comparison
use_rows <- filterByExpr(se, group=colData(se)[[params$group]], min.count = 5, min.total.count = 15, min.prop = 0.5)
se <- se[use_rows,]
```

TMM normalization
```{r dw-tmm-hs}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)
```

CPM calculation
```{r dw-cpm-hs}
assay(se, 'cpm') <- edgeR::cpm(assay(se, 'normcounts'))
assay(se, 'logcpm') <- edgeR::cpm(assay(se, 'normcounts'), log = TRUE)
```



### Additional objects
Selection of HVG
```{r conf-hvg}
# Variance modelling
fit_res <- modelGeneVar(se)
fit_md <- metadata(fit_res)
# Define highly variable genes
hvg <- getTopHVGs(fit_res, prop = 0.1)
hvg_top_n100 <- getTopHVGs(fit_res, n = 100)
```

Annotation variables and colors
```{r conf-annot}
ann_vars <- strsplit(params$ann_vars, ",") %>% unlist %>% 
  intersect(., colData(se) %>% colnames)
ann_df <- colData(se) %>% data.frame %>% 
  dplyr::select(all_of(ann_vars))

ann_colors <- foreach(i = ann_vars) %do% {
  get_palette(i)
}
names(ann_colors) <- ann_vars
ann_colors <- compact(ann_colors)
```

## PCA
```{r pca-conf}
use_rows <- hvg
mat <- assay(se[use_rows,], 'logcounts')
rownames(mat) <- make.names(rowData(se[use_rows,])$gene_name,unique = T)
# Create annotation, keeping only variables with more than 1 unique value
use_cols <- ann_df %>% dplyr::select(where(~n_distinct(.) > 1)) %>% colnames
p <- PCAtools::pca(mat, metadata = colData(se) %>% data.frame %>% dplyr::select(all_of(use_cols)))
```


### Determine optimum number of PCs to retain
Horn’s parallel analysis and elbow method
```{r pca-elbow, fig.width = 3.1, fig.asp = 0.9}
horn <- parallelPCA(mat)
elbow <- findElbowPoint(p$variance)
# horn$n
# elbow
screeplot(
  p,
  axisLabSize = 8, titleLabSize = 8,
  components = getComponents(p, 1:20),
  vline = c(horn$n, elbow)
) +
  geom_label(aes(x = horn$n + 0.5, y = 25, label = 'Horn\'s', vjust = -1),
             size = geom_text_size) +
  geom_label(aes(x = elbow + 0.5, y = 50, label = 'Elbow method', vjust = -1),
             size = geom_text_size)
```



### Variable correlation
Plot showing the correlation of PCs with sample-level variables. 
```{r pca-eigencor-plot, fig.width = 10.2}
use_rows <- hvg
n_pc <- seq_len(horn$n)
if(horn$n <= 2)
  n_pc <- 1:3

eigencor_res <- eigencorval(p, components = getComponents(p, n_pc), metavars = colnames(p$metadata))
# Select variables correlated with PC1 and PC2
col_vars <- colnames(eigencor_res$pvals)
col_pc12 <- c(
  col_vars[eigencor_res$pvals['PC1',]<0.05],
  col_vars[eigencor_res$pvals['PC2',]<0.05]
) %>% unique
```

```{r pca-eigencor-plot-curated, fig.width = 7.2}
use_rows <- hvg
n_pc <- seq_len(horn$n)
if(horn$n <= 2)
  n_pc <- 1:3

selected_vars <- c(
  `Anatomical origin` = 'anatomical_origin',
  `CTC type` = 'sample_type',
  `Number of cells` = 'n_cells',
  `Detected genes` = 'detected',
  `Total counts` = 'sum',
  `Mitochondrial %` = 'subsets_Mito_percent',
  `Ribosomal %` = 'subsets_Ribo_percent',
  `Mouse ID` = 'mouse_id',
  # `Cell cycle phase` = 'cc_phase_cy',
  `G1 score` = 'cc_g1_score_cy',
  `S score` = 'cc_s_score_cy',
  `G2M score` = 'cc_g2m_score_cy'
)

p_curated <- p  
p_curated$metadata <- p_curated$metadata %>% 
  dplyr::rename(all_of(selected_vars))
eigencor_res <- eigencorval(p_curated, components = getComponents(p_curated, n_pc), metavars = names(selected_vars))

# Plot correlation
eigencorplot(p_curated,
             components = getComponents(p_curated, n_pc),
             metavars = names(selected_vars),
             fontCorval = 2,
             cexCorval = 0.8,
             main = 'PC sample variables correlation')


```


### Bi-plot PC1-PC2
```{r pca-pc1-pc2-plot, fig.asp=0.8, fig.width = 2}
i <- 'donor'
use_palette <- ann_colors[[i]]
res <- PCAtools::biplot(
    p, 
    x = "PC1",
    y = "PC2",
    colby = i, 
    colkey = use_palette,
    colLegendTitle = 'donor',
    shape = params$group,
    shapekey = get_shape(params$group),
    shapeLegendTitle = 'Condition',

    legendPosition = 'right',
    pointSize = 1,
    lab = NULL, 

    ellipse = FALSE
  ) +
  xlim(c(-100, 50)) +
  ylim(c(-45, 45)) +
  theme_project

print(res) 

```

### Bi-plot PC2-PC3
```{r pca-pc2-pc3-plot, fig.asp=0.8, fig.width = 2}
i <- 'donor'
use_palette <- ann_colors[[i]]
res <- PCAtools::biplot(
    p, 
    x = "PC2",
    y = "PC3",
    colby = i, 
    colkey = use_palette,
    colLegendTitle = 'donor',
    shape = params$group,
    shapekey = get_shape(params$group),
    shapeLegendTitle = 'Condition',
    legendPosition = 'right',
    pointSize = 1,
    lab = NULL, 
    ellipse = FALSE
  ) +
  xlim(c(-100, 50)) +
  ylim(c(-45, 45)) +
  theme_project

print(res) 

```


